---
title: "Advanced Mapping"
layout: single
date: "2024-10-29"
output:
  pdf_document: default
  rmarkdown::html_vignette: default
toc_sticky: true
author_profile: true
toc: true
toc_label: Contents
vignette: "%\\VignetteIndexEntry{intermediate-mapping} %\\VignetteEngine{knitr::rmarkdown}
  %\\VignetteEncoding{UTF-8}\n"
---
```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

```

**Make sure** you've got the latest version of `ekonomR`. It's getting updated frequently. 

If you're not sure if your `ekonomR` is up to date or you're new to the woods, you may want to check out the vignette [Getting Started with ekonomR](https://stallman-j.github.io/ekonomR/vignettes/getting-started-with-ekonomR/), and the [Coding Review](https://stallman-j.github.io/ekonomR/vignettes/coding-review/) for a couple key operations I'll be assuming you know.

# Agenda

We're going to download a the basins data from [HydroATLAS](https://www.hydrosheds.org/hydroatlas) and plot all the basins and sub-basins for Africa, like the picture given [here](https://www.hydrosheds.org/products/hydrobasins)
 but just for Africa.
 
 
We'll then select a sub-basin that intersects with the Meru district in Tanzania, plot all levels of basins with that district, and export one of the sub-basins to a geojson file for later processin in Python.

# Prerequisites

First, bring `ekonomR` into your working library.

``` {r}
library(ekonomR)
```

# Set Parameters

Set the path that we want to download the data into. 

```{r, eval = TRUE, results = FALSE, echo = FALSE}
data_raw_path <- file.path("E:","data","01_raw")

```

```{r, eval = FALSE, results = FALSE}
#data_raw_path <- here::here("data","01_raw")
# if you want to save the data somewhere else
# data_raw_path <- file.path("E:","data","01_raw")

```


`locality_color` provides the HEX color codes for a medium Yale blue, which we'll use for mapping.

```{r}

  locality_color <- "#286dc0"
  hydrobasin_level <- 01
  equal_area_crs   <- "ESRI:102022"

```

# Bring in pre-existing data

The vignette [Intermediate Mapping](https://stallman-j.github.io/ekonomR/vignettes/intermediate-mapping/) already showed how to download a country shapefile from the [Global Administrative Areas (GADM)]((https://uwaterloo.ca/library/geospatial/collections/us-and-world-geospatial-data-resources/global-administrative-areas-gadm)). 

For the purposes of this vignette, we're going to look at the intersection of a district in Tanzania called Makiba. We want to find out which river basins intersect with it, using river basin data from [HydroSHEDS](https://www.hydrosheds.org/products/hydrobasins). 

We'll also look at all the basins and nested sub-basins in Africa, and plot those.

# Download data

If the path `here::here("data","01_raw","GADM")` doesn't already exist, the `download_data()` function from `ekonomR` will create the folder path for you. This `if` statement wraps around the call to download the data, so that if the file already exists, there's no need to re-download it.

We want to download a 2.5G file that's BasinATLAS, which requires a while. The download will time out if we don't increase the amount of time that R's trying to download something.

I've increased the timeout time to 3,000 seconds or 50 minutes. This is hopefully far too long for what you would need.

```{r}

options(timeout = 3000)

filename <- "BasinATLAS_Data_v10.gdb"

path <- file.path(data_raw_path,"HydroATLAS","BasinsATLAS",filename)
# only do this download and extract if the file doesn't already exist

if (!file.exists(path)) {
  
  ekonomR::download_data(data_subfolder = file.path("BasinsATLAS"),
                          data_raw = data_raw_path,
                          url = "https://figshare.com/ndownloader/files/20082137",
                         filename = filename,
                         zip_file = TRUE)
  }
```

Now we want to bring the data in. The file gets extracted to a *folder* called `"BasinATLAS_Data_v10.gdb"`, but the actual `gdb` file lives in trhe folder *inside* that.

Let's take a quick look at the layers. There are 1 through 10, and it looks like it's probably the case that the 941,012 multipolygons in level 10 are probably the most concentrated of the sub-basins. Let's bring in the first layer just to take a quick plot.

```{r}
in_path <- file.path(path,"BasinATLAS_v10.gdb")

sf::st_layers(in_path)
```

```{r}
tictoc::tic("Bring in the HydroBASIN shape files")
    hydro_basin <- sf::st_read(dsn = in_path,
                               layer = "BasinATLAS_v10_lev01")
tictoc::toc()

```
Make a quick plot

```{r}
  map <- ggplot2::ggplot() +
    ggplot2::geom_sf(data = hydro_basin,
          color = "gray70",
          fill = "gray99",
          alpha = 0.5,
          linewidth = .1) +
    # ggplot2::geom_sf(data = my_localities,
    #         alpha = .3,
    #         fill  = locality_color,
    #         color = locality_color,
    #         linewidth = .1) +
    # ggplot2::geom_text(data = my_divisions,
    #                    ggplot2::aes(X,Y, label = NAME_3),
    #                    size = 5) +
    ggplot2::labs(title = paste0("Basic Hydro Basins"),
         caption = c("Data from HydroATLAS (2024)")) +
    ekonomR::theme_minimal_map(axis_title_x = ggplot2::element_blank(),
                               axis_title_y = ggplot2::element_blank())

#map
```

Output

```{r}
out_path <- file.path("C:","Projects","water-in-africa","output","03_maps")


  ekonomR::ggsave_map(output_folder = out_path,
           plotname = map,
           filename = paste0("HydroBASINS_lev01.png"),
           width = 9,
           height = 6,
           dpi  = 300)
```

# Keep basins in Africa

We have this basin data and we'd like to keep only the basins that intersect with Africa. 
`ekonomR` loads the package `rnaturalearth` which has some simple continent maps that we can use.

We don't care at present about the country boundaries, so we take `sf::st_union()` to aggregate them. We just want a simple intersection, so for right now we'll set spherical geometry not to use the s2 spherical geometry package.

```{r}
sf::sf_use_s2(FALSE)
continent_polygon <- rnaturalearth::ne_countries(continent = "africa") %>%
                    sf::st_make_valid() %>%
                    sf::st_union()

```
Do a quick plot to check that looks about right:

```{r}
plot(sf::st_geometry(continent_polygon))
```

# Get intersection of Africa polygon and HydroBASINS

```{r}

intersected_basins <- sf::st_intersection(hydro_basin,continent_polygon)
```
Plot it out to see that this makes sense. Looks like two main basins, Africa mainland and Madagascar.

```{r}
plot(sf::st_geometry(intersected_basins))

```

# Get next-level-down basins and intersect with Africa

Now let's get the basins of the next level down, 02. For Africa, there are 10 of these out of 62 globally.

```{r}
tictoc::tic("Bring in the HydroBASIN shape files")
    basin_02 <- sf::st_read(dsn = in_path,
                               layer = "BasinATLAS_v10_lev02")
tictoc::toc()
```
Intersect again with Africa...

```{r}

intersected_basins_02 <- sf::st_intersection(basin_02,continent_polygon)
```
Plot to double-check:

```{r}
plot(sf::st_geometry(intersected_basins_02))

```

# Sub- Basins

Do it all together for the next levels down.

At level 3 basins, we have 292 sub-basins globally, and 42 in Africa.

```{r}
tictoc::tic("Bring in the HydroBASIN shape files")
    basin_03 <- sf::st_read(dsn = in_path,
                               layer = "BasinATLAS_v10_lev03")
tictoc::toc()


intersected_basins_03 <- sf::st_intersection(basin_03,continent_polygon)

plot(sf::st_geometry(intersected_basins_03))


```

At level 4, we have 1,342 sub-basins globally, 244 of which are in Africa.


```{r}
tictoc::tic("Bring in the HydroBASIN shape files")
    basin_04 <- sf::st_read(dsn = in_path,
                               layer = "BasinATLAS_v10_lev04")
tictoc::toc()


intersected_basins_04 <- sf::st_intersection(basin_04,continent_polygon)

plot(sf::st_geometry(intersected_basins_04))


```
At level 5, there are 4,734 basins globally, and 1,020 in Africa.

```{r}
tictoc::tic("Bring in the HydroBASIN shape files")
    basin_05 <- sf::st_read(dsn = in_path,
                               layer = "BasinATLAS_v10_lev05")
tictoc::toc()


intersected_basins_05 <- sf::st_intersection(basin_05,continent_polygon)

plot(sf::st_geometry(intersected_basins_05))


```
At level 6, there are 16,397 sub-basins globally, and 3,591 in Africa.

```{r}
tictoc::tic("Bring in the HydroBASIN shape files")
    basin_06 <- sf::st_read(dsn = in_path,
                               layer = "BasinATLAS_v10_lev06")
tictoc::toc()


intersected_basins_06 <- sf::st_intersection(basin_06,continent_polygon)

plot(sf::st_geometry(intersected_basins_06))


```

At level 7, there are 57,646 sub-basins globally, and 12,350 in Africa.

```{r}
tictoc::tic("Bring in the HydroBASIN shape files")
    basin_07 <- sf::st_read(dsn = in_path,
                               layer = "BasinATLAS_v10_lev07")
tictoc::toc()


intersected_basins_07 <- sf::st_intersection(basin_07,continent_polygon)

plot(sf::st_geometry(intersected_basins_07))


```
```{r}
tictoc::tic("Bring in the HydroBASIN shape files")
    basin_08 <- sf::st_read(dsn = in_path,
                               layer = "BasinATLAS_v10_lev08")
tictoc::toc()


intersected_basins_08 <- sf::st_intersection(basin_08,continent_polygon)


```


```{r}
tictoc::tic("Bring in the HydroBASIN shape files")
    basin_09 <- sf::st_read(dsn = in_path,
                               layer = "BasinATLAS_v10_lev09")
tictoc::toc()


intersected_basins_09 <- sf::st_intersection(basin_09,continent_polygon)


```


If we go all the way to level 10, we get 941,012 sub-basins globally, and 209,367 in Africa.

```{r}
tictoc::tic("Bring in the HydroBASIN shape files")
    basin_10 <- sf::st_read(dsn = in_path,
                               layer = "BasinATLAS_v10_lev10")
tictoc::toc()


intersected_basins_10 <- sf::st_intersection(basin_10,continent_polygon)


```


```{r, results = "hide"}
  map <- ggplot2::ggplot() +
  ggplot2::geom_sf(data = intersected_basins_06,
            alpha = .3,
            fill  = "gray99",
            color = locality_color,
            linewidth = .1) +
    ggplot2::geom_sf(data = intersected_basins_03,
            alpha = .3,
            fill  = "gray99",
            color = "black",
            linewidth = .5) +
    ggplot2::labs(title = paste0("Africa, Level 3 and Level 6 Sub-Basins"),
         caption = paste0("Data from HydroSHEDS (2013) \n Includes ",nrow(intersected_basins_03), " level 3 and ",nrow(intersected_basins_06), " level 6 basins.")) +
    ekonomR::theme_minimal_map(axis_title_x = ggplot2::element_blank(),
                               axis_title_y = ggplot2::element_blank())

map
```

And save

```{r}
out_path <- file.path("C:","Projects","water-in-africa","output","03_maps")


  ekonomR::ggsave_map(output_folder = out_path,
           plotname = map,
           filename = paste0("Africa_hydrobasins_level03_level06.png"),
           width = 9,
           height = 8,
           dpi  = 300)
```    

# Get intersection of Level 7 Basin with a Division in Tanzania

There's a boundaries file in `ekonomR` that gives the shapefile from GADM of Makiba, one of Tanzania's districts in which gravity irrigation has taken place for at least a century.

We'll bring it in, and then calculate which of the Level 7 Basins intersects with it, then export that basin to a geojson.

```{r}
data("makiba")

makiba_ea <- makiba %>%
             sf::st_transform(crs = equal_area_crs)

makiba_ea <- cbind(makiba_ea,
                       sf::st_coordinates(sf::st_centroid(makiba_ea)))

basin_09_ea <- intersected_basins_09 %>%
             sf::st_transform(crs = equal_area_crs)

basin_09_ea <- cbind(basin_09_ea,
                       sf::st_coordinates(sf::st_centroid(basin_09_ea)))
              

```

Get intersection. This is going to be more problematic at a smaller scale, because the basins and Makiba are both small. We might prefer to choose an equal-area CRS and so that we're taking planar calculations.



```{r}

#intersected_basin_makiba <- sf::st_intersects(basin_10_ea,makiba_ea)

does_intersect = lengths(sf::st_intersects(basin_09_ea,makiba_ea))>0

my_basin <- basin_09_ea[does_intersect,]

plot(sf::st_geometry(my_basin))
```



```{r}
  map1 <- ggplot2::ggplot() +
  ggplot2::geom_sf(data = my_basin,
            alpha = .3,
            fill  = "gray99",
            color = locality_color,
            linewidth = .1) +
        ggplot2::geom_sf(data = makiba_ea,
            alpha = 0,
            fill  = "gray70",
            color = "black",
            linewidth = .5) +
      ggplot2::geom_text(data = makiba_ea,
                       ggplot2::aes(X,Y, label = NAME_3),
                       size = 5) +
    ggplot2::labs(title = paste0("Makiba, Tanzania and its associated Level 9 Basins"),
         caption = paste0("Data from HydroSHEDS (2013).")) +
    ekonomR::theme_minimal_map(axis_title_x = ggplot2::element_blank(),
                               axis_title_y = ggplot2::element_blank())

map1

# save map
  ekonomR::ggsave_map(output_folder = out_path,
           plotname = map1,
           filename = paste0("Makiba_level-09_basins.png"),
           width = 8,
           height = 8,
           dpi  = 300)

```
```{r}
  country        <- "TZA"
  country_name   <- "Tanzania"
  locality_color <- "#286dc0"
  base_url       <- "https://geodata.ucdavis.edu/gadm/gadm4.1/gpkg/"
  gadm_level     <- 1
  filename       <- paste0("gadm41_",country,".gpkg")
  in_path        <- file.path(data_raw_path,"GADM",filename)
  name_var       <- "NAME_2"
  num_localities <- 1
  names_to_keep  <- c("Meru")
```

# Download data

If the path `here::here("data","01_raw","GADM")` doesn't already exist, the `download_data()` function from `ekonomR` will create the folder path for you. This `if` statement wraps around the call to download the data, so that if the file already exists, there's no need to re-download it.

```{r}

if (!file.exists(file.path(data_raw_path,"GADM",filename))) {
  ekonomR::download_data(data_subfolder = "GADM",
                          data_raw = data_raw_path,
                          url = paste0(base_url,filename),
                          filename = filename)
}
```

Read in the country-level shapefile. To make taking the intersection easier, we use level 1 so that there are fewer rows to deal with

```{r}
country_gpgk <- sf::st_read(dsn = in_path,
                            layer = paste0("ADM_ADM_",gadm_level)) %>%
                sf::st_make_valid()

```
Then take the union of all the level-1 subdivisions. We'll start with taking any basin that intersects with 
