---
title: "Simple Plotting with a Review of R Coding Basics"
layout: single
toc: true
toc_label: "Contents"
toc_sticky: true
author_profile: true
date: "2024-09-23"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{basic-plotting}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Getting Started

If you're working through this vignette with an eye towards starting your own project, I *highly* recommend first checking out the [package documentation](https://stallman-j.github.io/ekonomR/documentation/documentation/) and the vignette [Getting Started with ekonomR](https://stallman-j.github.io/ekonomR/vignettes/getting-started-with-ekonomR/) to get your project structured in a way that's scalable, sharable, and documentable. 

You can check out the full list of vignettes [here](https://stallman-j.github.io/ekonomR/vignettes/vignettes/)

If you've already created a project using the `ekonomR` function `create_folders()`, you may want to copy the code from the end of this vignette into a file called, say, `basic-plotting-vignette.R` into your project folder folder `code/scratch` so that you can edit and refer back to it.

If you're familiar with RMarkdown or you'd like an excuse to learn it, you can copy `basic-plotting.Rmd` from [the GitHub repo for ekonomR](https://github.com/stallman-j/ekonomR/blob/main/vignettes/basic-plotting.Rmd) and save it into `code/scratch`.

Exercises called *comprehension check* will be those that you may understand just by looking at the code if you're experienced in R. If it's not obvious to you how you would write the code to answer these checks, you should puzzle around with the code in your console to figure them out.

There's a more involved exercise at the very end that you can build out on your own in an R script by modifying the code in here.

We'll be plotting country-level emissions data from the Global Carbon Budget, which you can learn about [here](https://globalcarbonbudget.org/). (To add: Proper data citation)

The data we'll use has been cleaned and loaded into the package `ekonomR`. 

## New Installation 

Run these two lines in your console. This will allow the updated version of `ekonomR` to get installed into your library. If you've already installed the R package `remotes`, comment out that line with a `#` sign in front.

``` {r install, results = FALSE, eval = FALSE}
install.packages("remotes") 
remotes::install_github("stallman-j/ekonomR")
```

## Re-installation 

If you've already installed `ekonomR` before starting this vignette, you'll need to re-install it correctly so that you can access this update.

First, go into the "Packages" tab in RStudio (it's in the window that's shared with tabs for `Files`, `Packages`, `Help`, `Viewer`, and `Presentation`) and make sure that `ekonomR` is *unchecked*. If you don't do this, you might get an error message or R will have to restart.

Then run these two lines in your console. This will allow the updated version of `ekonomR` to get installed into your library. If you've already installed the R package `remotes`, comment out that line with a `#` sign in front.

``` {r, results = FALSE, eval = FALSE}
install.packages("remotes") 
remotes::install_github("stallman-j/ekonomR")
```

Either way, once you've installed `ekonomR`, you'll want to bring the `ekonomR` package into your working library.

```{r setup}
library(ekonomR)
```

Your R Session might ask you to download a bunch of packages. This isn't usually a problem, but because `ekonomR` is getting updated so frequently, you might run into trouble.

If you're given the option, update packages from CRAN, the package repository for well-documented R packages. If your R crashes, run the above sequence but then instruct R *not* to update the packages and see how things go. If you're still having trouble, try uninstalling and reinstalling R and R Studio and then coming back. If you're still having trouble, email me.

Why is `ekonomR` doing this? 

  - `ekonomR` is installing and loading most of the packages that you'll have to use anyways for an economic analysis. These are its **dependencies**: the packages and other files `ekonomR` itself relies on to run properly.
  - Hopefully we're *not* installing a bunch of the packages you won't need, though.

# Data Exploration

Let's tell R that we want to use the cleaned GCB data.

```{r}
data(gcb)
```


We're going to plot territorial emissions for China for all the available years. This is a **time series**: we'll be showing the change in a single unit (here, a country), over time.

We won't go deep into data exploration for the purposes of this vignette, but another vignette will be out about the cleaning of this data and we'll explore some more in that one. R calls most data a data frame, which you can think of as a single sheet in an Excel workbook. 

We need a very basic understanding of what the data look like, however.

In your RStudio console, input the following:

```{r}
View(gcb)
names(gcb)
```

*Comprehension check:* What is `names(gcb)` giving us for output? What class is this object? (Hint: `class(names(gcb))`.) What is the class of `gcb`?

Let's see what years we have available (Note: Not all years will be available for all measures)

```{r}
unique(gcb$year)
```


*Comprehension Exercise:* How many unique countries are present in this data frame?



# Coding Review


It's good coding practice to *soft-code* wherever you can. This means that rather than inputting "China" everywhere, and then replacing it everywhere if we later want to plot something else, let's try to define at the top of our scripts something that we only have to change up here and not throughout.

```{r}
chosen_country <- c("CHN")
chosen_country_name <- "China"
```

These will show up in other code blocks throughout this vignette. 

Let's also create a data frame that just contains the data for China.

```{r}
data_country <- gcb %>% 
                dplyr::filter(iso3c == chosen_country)
```


## The Pipe Operator %>%

If you're not familiar with the data cleaning and organizing packages included in what's called the `tidyverse`, you might be a little confused by the `%>%` symbol. This is called the "pipe operator." Think of it as a funnel.

You take the stuff that came before the pipe, and funnel it through the pipe into the next function that comes below.

In coding terms, what it says is, take the thing that came before (here `gcb`) and insert it into the first argument of the function that comes next (here, `filter`). 

We could also have written the following: 

```{r}
data_country <- dplyr::filter(gcb, iso3c == chosen_country)
```

*Comprehension check:* What has the `filter` function done? How many rows exist in the data frame `data_country`?

Writing `dplyr::filter` just tells R that the function `filter` comes with the `dplyr` package. There are several packages that have a `filter` function, so we specify *which* package's `filter` function we want here. 

It's good practice to state the package that you're getting functions from. When someone else uses your code, or you use your own code on a different machine, or you update R, this will signal to R that it should install the latest version of those packages.

Look at the help for this function
```{r}
?filter
```

and click on the `filter` function for `dplyr`.

*Comprehension Check*: Did you see another package that also has a `filter` function? If so, which was it?

You'll see that the first thing that goes into the function is `.data`. `filter` takes a data frame, and keeps only the *rows* for which a condition holds (here, that the iso3c code is the same as we've listed for `chosen_country`.)

The reason for using `%>%` is that there may be many data manipulation operations we'd like to do, and once you get used to it, piping makes your code pretty simple to write and very simple to follow. Here, it's a little unnecessary because we just want one thing to happen: we want to keep only the observations for China.

**Nerd tip:** If you're used to SQL or interested in big data, you should be thinking hard about using the package `data.table` rather than the `tidyverse` packages including `dplyr`. `data.table` is just far, far faster than `dplyr` can manage once your data get big.

## == or =?

The double equals sign, `==`,  is not the same as the single equals sign, `=`. 

The double equals sign is a logical check: for certain elements, R is checking whether the condition on the left (here, the element's `iso3c`) is equal to the thing on the right (here, that it equals `chosen_country`, which we defined above to be `"CHN"`). 

A single equals sign would be setting the left-hand side equal to the right-hand side. There's a place that would have made sense to use this: We could have written `chosen_country = c("CHN")` to set the value for `chosen_country`. 

In R it's common to use the left-facing arrow `<-` in your scripts to make that sort of assignment. What a line with this arrow assignment means is: "Set the thing on the right to be equal to the thing that the arrow is pointing to". However, the equals sign, `=`, is commonly used for this sort of assignment in functions, and it's common to see it used in regular script by people who come to R from other languages like Python.

In our context, `iso3c == "CHN"` will return `TRUE` if the row is for China, and `FALSE` if the observation is not for China. 


## Filtering with Multiple Conditions

If you click on this data frame in your Environment tab in RStudio, you may notice a number of empty cells: these are missing observations. 

Since we want to plot territorial emissions, we can add a condition to the filter so that we can ignore those missing values:

```{r}
data_country <- gcb %>% dplyr::filter(iso3c == chosen_country & !is.na(gcb_ghg_territorial))
```

We've added a condition that has to be true: we are now removing the rows for which `gcb_ghg_territorial` is missing. We've also *replaced* the old `data_country` with our new one.


## Logical (Boolean) Statements

A **logical** or **Boolean** statement is one which is either `TRUE` or `FALSE` (commonly also coded as `1` or `0`).

`is.na()` is a function that asks a logical question: it creates a vector of the same length as the vector you feed it. In its output vector, it will return `TRUE` for the ith element of the vector it is examining if the spot of the ith element is missing (`NA`).

The `!` is a negation, so `!is.na()` now returns `TRUE` if the element is *not* missing. This means in our context that we keep the rows for which the territorial emissions *are* present.

It might be a little opaque what vector `!is.na()` is examining, but it's the column vector called `gcb_ghg_territorial` in the data frame `gcb`.

We could write that vector as `gcb$gcb_ghg_territorial`. 

Using `dplyr::filter` masks that a little bit because we already stated up at the beginning that we're looking inside the data frame `gcb`.

## Multiple Pipes

Let's see the `%>%` in action with a little more complexity. Since we're just plotting territorial emissions, let's pick out (select) only the relevant columns.

```{r}
data_country <- gcb %>% 
                dplyr::filter(iso3c == chosen_country & !is.na(gcb_ghg_territorial)) %>%
                dplyr::select(year,country_name,iso3c,gcb_ghg_territorial)
```

**Note:** if you're using the pipe, you have to make sure that it goes at the end of the line, not the beginning.

This code block is saying the following:

1. First take the data frame `gcb` (that's what goes before the first pipe). 
2. Then, take only the rows where the `iso3c` is equal to `CHN` and also the `gcb_ghg_territorial` is not missing (the second line, before the second pipe)
3. Once you've done that, keep only the columns `year`, `country_name`, `iso3c`, and `gcb_ghg_territorial`. 

There are many ways we could write a script in R to get to this result. Just with the functions we've seen, we could split the filter into two separate pipes. We could swap the order of the `filter` and `select`. 

In this case the order of what goes into the pipes won't matter. But in other cases, with more complicated cleaning, the order that you do these operations may well make a difference.

*Comprehension check:* Create a new dataframe called `data_country2` that's exactly the same as `data_country` but uses a slightly different way of writing the command to get there from `gcb`.

# Plotting with ggplot2

Now let's get to plotting. The package `ggplot2` is a versatile plotting package that allows you to use a similar syntax for plotting all sorts of figures, from bar charts to complicated maps. 

There's a [whole ggplot2 book](https://ggplot2-book.org/) that you can use to get into the details, but I've found that I mostly end up tinkering with a few of these operations, and that's why `ekonomR` has a function called `theme_minimal_plot()` which wraps around the `ggplot2` functions `theme_minimal()` and `theme()` to give us a nicely formatted plot, in a good size for putting in a paper or a presentation, that we don't have to worry too much about.

The way `ggplot2` typically works is that we start with an empty plot, and then we add visual components in layers with each new call to something of the form `geom_xxx`, or we add on options about the captions or titles or visual elements.

Let's build up our plot. The skeleton of it is the following:

```{r, results = "hide"}
my_plot <- ggplot2::ggplot() + 
  ggplot2::geom_point(data = data_country,
                      ggplot2::aes(x = year, 
                                   y =gcb_ghg_territorial)
                      )
```
```{r, echo = FALSE, results = "hide"}
ggsave_plot(output_folder = here::here("output","02_figures"),
         plotname = my_plot,
         filename = paste0("gcb_territorial_emissions_",chosen_country_name,"_plot-01.png"),
         width = 8,
         height = 6,
         dpi  = 400)
```

```{r, echo = FALSE}
url <- "https://github.com/stallman-j/ekonomR/blob/main/output/02_figures/gcb_territorial_emissions_China_plot-01.png?raw=true"
knitr::include_graphics(url)

```
You should input `my_plot` into the console of RStudio to see if the plot shows up in your plotting pane like you would expect. Try doing this after each instance that we change something in our plots.

`ggplot2::ggplot()` just opens up a new, blank plot for us. Try putting just this in your console and see what happens.

Under this, we add (with the plus sign) the following: `ggplot2::geom_point()`, which calls a scatter plot. The scatter plot uses the data from `data_country`, and sets as the x axis its variable `year` and its y axis the variable `gcb_ghg_territorial`.

We can make this a little neater. Let's change the axis labels and add both a caption to cite our data and a title. We add with another addition sign these options below our original plotting code.


```{r, results = "hide"}
my_plot <- ggplot2::ggplot() + 
  ggplot2::geom_point(data = data_country,
                      ggplot2::aes(x = year, 
                                   y =gcb_ghg_territorial)
                      ) +
  ggplot2::labs(title = paste0("Territorial Emissions, ",chosen_country_name),
       caption = c("GDP from GCB (2023)"),
       x ="" ,
       y = "Emissions (units here)"
  )

```
```{r, echo = FALSE, results = "hide"}
ggsave_plot(output_folder = here::here("output","02_figures"),
         plotname = my_plot,
         filename = paste0("gcb_territorial_emissions_",chosen_country_name,"_plot-02.png"),
         width = 8,
         height = 6,
         dpi  = 400)
```

```{r, echo = FALSE}
url <- "https://github.com/stallman-j/ekonomR/blob/main/output/02_figures/gcb_territorial_emissions_China_plot-02.png?raw=true"
knitr::include_graphics(url)

```

Better! We're getting somewhere. But the grid lines are a little odd and why is the background grey? 

`ggplot2` has a bunch of themes. A nice one is `theme_minimal()`, which you can add on:
```{r, results = "hide"}
my_plot <- ggplot2::ggplot() + 
  ggplot2::geom_point(data = data_country,
                      ggplot2::aes(x = year, 
                                   y =gcb_ghg_territorial)
                      ) +
  ggplot2::labs(title = paste0("Territorial Emissions, ",chosen_country_name),
       caption = c("GDP from GCB (2023)"),
       x ="" ,
       y = "Emissions (units here)"
  ) + 
  ggplot2::theme_minimal()


```

```{r, echo = FALSE, results = "hide"}
ggsave_plot(output_folder = here::here("output","02_figures"),
         plotname = my_plot,
         filename = paste0("gcb_territorial_emissions_",chosen_country_name,"_plot-03.png"),
         width = 8,
         height = 6,
         dpi  = 400)
```

```{r, echo = FALSE}
url <- "https://github.com/stallman-j/ekonomR/blob/main/output/02_figures/gcb_territorial_emissions_China_plot-03.png?raw=true"
knitr::include_graphics(url)

```
Still not loving the background.

We could keep doing this, finagling with elements. It was through tinkering with this process that I realized there are some settings I would like to just *have* without having to go look up. 

That's why `ekonomR` has a function that adds a little bit onto `ggplot2`'s minimal theme to preserve those defaults. It's called `theme_minimal_plot()`.

If you just want a nice, black-and-white plot that is pretty simple to change the labels on, just tack it on like so:

```{r, results = "hide"}
my_plot <- ggplot2::ggplot() + 
  ggplot2::geom_point(data = data_country,
                      ggplot2::aes(x = year, 
                                   y =gcb_ghg_territorial)
                      ) +
  ggplot2::labs(title = paste0("Territorial Emissions, ",chosen_country_name),
       caption = c("GDP from GCB (2023)"),
       x ="" ,
       y = "Emissions (units here)"
  ) + 
  ekonomR::theme_minimal_plot()

```
```{r, echo = FALSE, results = "hide"}
ggsave_plot(output_folder = here::here("output","02_figures"),
         plotname = my_plot,
         filename = paste0("gcb_territorial_emissions_",chosen_country_name,"_plot-04.png"),
         width = 8,
         height = 6,
         dpi  = 400)
```

```{r, echo = FALSE}
url <- "https://github.com/stallman-j/ekonomR/blob/main/output/02_figures/gcb_territorial_emissions_China_plot-04.png?raw=true"
knitr::include_graphics(url)

```

That'll do, right? This plot is nice and crisp. The contrast is fairly high. The sizing is pretty good on titles and captions. We could keep tinkering, but the marginal returns are pretty low at this point. You can see more options I often end up changing with `?theme_minimal_plot` if you like.


Now let's save the plot, with a function from `ekonomR` slightly adapting ggplot2's `ggsave` with some useful defaults.


```{r, results = "hide"}
ekonomR::ggsave_plot(output_folder = here::here("output","02_figures"),
         plotname = my_plot,
         filename = paste0("gcb_territorial_emissions_",chosen_country_name,".png"),
         width = 8,
         height = 6,
         dpi  = 400)
```

You can see this final plot by typing `my_plot` in the console, but you should check and make sure that the PNG lives where you can find it.

The `here::here()` function is amazing. Try typing it in your console. It should output a file path for you. This function cleverly looks for where it thinks your home project folder is, and then defines directories relative to this folder. If you've got this code running in an R Project (which you should), then your output from `here::here()` should be that project folder's path.

We defined `chosen_country_name` way up at the top. It's also coming in down here with the `paste0` function, which is also a very useful function. 

This means that if we went up to the top and changed *just* `chosen_country` and `chosen_country_name`, we could run through all this code without changing anything else, and get a plot for Chile or Kenya or the United States rather than China. That's cool!

*Comprehension Check:* What is `paste0` doing in the above code block? Plug `paste0("gcb_territorial_emissions_",chosen_country_name,".png")` into your console and see what gets outputted. If you're not sure, try making up something like `paste0("ekonomR ","is really really nerdy", "... ","but fun..."," in a very nerdy way")` and then toy around a bit with it.

# Exercises

For a country which is  *not* China, plot its greenhouse gas *consumption* (not territorial!) emissions over time with the following changes to your final product.

- Edit the y axis labels to contain the appropriate units. They currently say `Emissions (units here)`.
    -*Hint:* You may want to poke around in the [GCB page](https://globalcarbonbudget.org/) (the data page of which is [here](https://globalcarbonbudgetdata.org/latest-data.html) to figure out what the units should be.
    
- Correct the caption to contain the correct data attribution. It should be something like "Emissions data from GCB (2023)."
- Change the title to reflect that you're now plotting emissions from consumption. You should also make sure that the country you're listing in the title is the country that you show data for (note that we soft-coded that!).

- Think about what you see from these trends relative to what you know about the history of the country.

  - Do you think there's evidence for a Green Kuznets curve for this country? 
  - Do these trends accord with your priors?
  - What else would you need to examine in order to make a more definitive claim about a relationship between greenhouse gas emissions and the growth trajectory of this country?

- What quibbles might you have with the data itself? 

  - Do you trust the exact numbers? If not, do you think it's getting the general direction right? 
  - If there are biases in the data, what direction do you think they would take?


# Up Next: Intermediate Plotting

See the next vignette on [Intermediate Plotting](https://stallman-j.github.io/ekonomR/vignettes/intermediate-plotting/) for more plotting examples including layered plots and finer-tuned settings.

(Coming soon)

# Just the code

Here's just the code from the vignette in case you want to copy it into your own script. This is highly recommended for doing the exercise. 

You may want to add in comments so you know what's happening, though. (Hint: You can comment out a single line or part of a line with a `#`. If you want to comment out multiple lines at once, in RStudio you can first highlight the lines you want to comment out, and then use the keyboard command `Ctrl`+`Shift`+`C`. This will put a pound sign before every line.

``` {r, results = "hide", eval = FALSE}

# Setup
install.packages("remotes") 
remotes::install_github("stallman-j/ekonomR")
library(ekonomR)

# Data Exploration
data(gcb)
View(gcb)

names(gcb)

unique(gcb$year)

# Coding Review
chosen_country <- c("CHN")

chosen_country_name <- "China"

data_country <- gcb %>% 
                dplyr::filter(iso3c == chosen_country)
                
#data_country <- dplyr::filter(gcb, iso3c == chosen_country)

data_country <- gcb %>% dplyr::filter(iso3c == chosen_country & !is.na(gcb_ghg_territorial))

#data_country <- gcb %>% 
#                dplyr::filter(iso3c == chosen_country & !is.na(gcb_ghg_territorial)) %>%
#                dplyr::select(year,country_name,iso3c,gcb_ghg_territorial)

years_to_show <- c(1958,1959,1960,1961,1978,1995,2000,2015,2019)

# Plotting

my_plot <- ggplot2::ggplot() + 
  ggplot2::geom_point(data = data_country,
                      ggplot2::aes(x = year, 
                                   y =gcb_ghg_territorial)
                      ) +
  ggplot2::labs(title = paste0("Territorial Emissions, ",chosen_country_name),
       caption = c("GDP from GCB (2023)"),
       x ="" ,
       y = "Emissions (units here)",
       color = "Emissions" # sets legend name
  ) + 
  ekonomR::theme_minimal_plot()

my_plot

ggsave_plot(output_folder = here::here("output","02_figures"),
         plotname = my_plot,
         filename = paste0("gcb_territorial_emissions_",chosen_country_name,".png"),
         width = 8,
         height = 6,
         dpi  = 400)
```
