---
title: "Basic Cleaning: Penn World Tables"
layout: single
toc: true
toc_label: "Contents"
toc_sticky: true
author_profile: true
date: "2024-10-23"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{basic-cleaning_pwt}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

library(dplyr)
```


**Make sure** you've got the latest version of `ekonomR`. It's getting updated frequently. 

If you're not sure if your `ekonomR` is up to date or you're new to the woods, you may want to check out the vignette [Getting Started with ekonomR](https://stallman-j.github.io/ekonomR/vignettes/getting-started-with-ekonomR/), and the [Coding Review](https://stallman-j.github.io/ekonomR/vignettes/coding-review/) for a couple key operations I'll be assuming you know.

# Prerequisites

In this vignette, I'm going to assume that you already know everything we covered in [Basic Cleaning: Global Carbon Budget](https://stallman-j.github.io/ekonomR/vignettes/basic-cleaning_gcb/), so you might want to go there first. 

If you're just looking for a script that cleans the Penn World Tables because that's a very common GDP measure, you've also come to the right place and you might want to just copy the final code at the end.

First, bring `ekonomR` into your working library.

``` {r}
library(ekonomR)
```

# Download

In this vignette, we're going to download and clean the latest version of the [Penn World Tables](https://www.rug.nl/ggdc/productivity/pwt/?lang=en) (PWT). The PWT are perhaps the most highly regarded source for data on gross domestic product among economists. 

Let's go ahead and download the data. `ekonomR` has a `download_data()` function that allows for pretty easy downloading and decompressing of common formats that we would get data for.

``` {r}

  url <- "https://dataverse.nl/api/access/datafile/354095"

  ekonomR::download_data(data_subfolder = "PWT",
                data_raw       = here::here("data","01_raw"),
                url            = url,
                filename       = "pwt1001.xlsx")
```

You can look in the file path given by `here::here("data","01_raw","PWT")` to see where the excel file has gone. In it, there's a sheet called "Data" that we should bring in.


```{r}

  pwt <- readxl::read_xlsx(path = here::here("data","01_raw","PWT","pwt1001.xlsx"),
                     sheet = "Data",
                     col_names = TRUE)
```

If you check the [Penn World Tables page](https://www.rug.nl/ggdc/productivity/pwt/?lang=en), you'll see that there are a whole bunch of measures that all, ostensibly, could be considered as gross domestic product (GDP). 

For this particular project, we ultimately want to look at real GDP. This vignette series is loosely based around assessing the hypothesis that as countries grow richer, their environmental quality first worsens, and then improves. This is what's known as the hypothesis of the Environmental Kuznets Curve.

However, since we're looking over decades, we want *real* GDP which takes out the impact of inflation, and allows us to compare the living standards across countries, and over time.

The measure that comes closest to that is described as RGDPe, and it's inputted in the PWT in millions of 2017 US dollars. 

To keep things simple, we'll just keep the variable names that we want for this particular analysis. Let's also rename the variable `countrycode` to be `iso3c`. That'll allow us to merge with the emissions data more easily.

Putting this all together:

``` {r}
  names(pwt)

  pwt_clean <- pwt %>%
               dplyr::rename(iso3c = countrycode) %>%
               dplyr::select(iso3c, country, year, rgdpe)
```

Finally, let's save it all with the function `save_rds_csv()` from `ekonomR` to get both an excel version and an RDS version of this file.

```{r}
  pwt_clean <- ekonomR::save_rds_csv(data = pwt_clean,
                          output_path   = here::here("data","03_clean","PWT"),
                          output_filename = paste0("pwt_clean.rds"),
                          remove = FALSE,
                          csv_vars = names(pwt_clean),
                          format   = "xlsx")

```

That's all there is to it! 

This is one way you can see that the PWT are used a lot by economists: we could pretty much load this dataset into R and start doing analysis with it right away. The code required for cleaning is very minimal.

# Just the code, please

```{r, eval = FALSE}
# bring in ekonomR to the library
library(ekonomR)

# Download the data

  url <- "https://dataverse.nl/api/access/datafile/354095"

  ekonomR::download_data(data_subfolder = "PWT",
                data_raw       = here::here("data","01_raw"),
                url            = url,
                filename       = "pwt1001.xlsx")
  
# Read in the data
  
  pwt <- readxl::read_xlsx(path = here::here("data","01_raw","PWT","pwt1001.xlsx"),
                     sheet = "Data",
                     col_names = TRUE)
  
# Choose our variables of interest
  
  pwt_clean <- pwt %>%
               dplyr::rename(iso3c = countrycode) %>%
               dplyr::select(iso3c, country, year, rgdpe)
  
# Save the cleaned data

  pwt_clean <- ekonomR::save_rds_csv(data = pwt_clean,
                          output_path   = here::here("data","03_clean","PWT"),
                          output_filename = paste0("pwt_clean.rds"),
                          remove = FALSE,
                          csv_vars = names(pwt_clean),
                          format   = "xlsx")
  
```
