---
title: "Basic Plotting"
layout: single
toc: true
toc_label: "Contents"
toc_sticky: true
author_profile: true
date: "2024-09-23"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{basic-plotting}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Getting Started

Throughout, exercises called *comprehension check* will be those that you may understand just by looking at the code if you're experienced in R. If these checks aren't obvious to you, you should implement the code to figure them out.

There's a more involved exercise at the very end that you're encouraged to build out on your own in an R script. If you're working with the `ekonomR` sequence and you've created a project 

# Bring in the data

We'll be plotting data from the Global Carbon Budget, which you can learn about [here](https://globalcarbonbudget.org/). 



This data has been cleaned and loaded into the package `ekonomR`. If you've already installed `ekonomR`, you'll want to go into the "Packages" tab and make sure that `ekonomR` is *unchecked* before you run the next two lines in your console. This will allow the updated version of `ekonomR` to get installed into your library.

``` {r install}
install.packages("remotes")
remotes::install_github("stallman-j/ekonomR")
```

```{r setup}
library(ekonomR)
```

Let's tell R that we want to use this data.

```{r}
data(gcb_clean)
```

Let's also remove the scientific notation in our axes. Otherwise we get x and y axis as e^4 and this gets difficult to read. We'd rather take logs in most cases, or rescale the axes, instead of dealing with scientific notation in our plots.

```{r}
options(scipen = 999)
```
(Future vignettes will go through the details of downloading and cleaning this data).

# Data Exploration

Let's look at this data. R calls it a data frame, which you can think of as a single sheet in an Excel workbook. 

In your console, enter the following:

```{r}
View(gcb_clean)
names(gcb_clean)
```

*Comprehension check:* What is `names(gcb_clean)` giving us for output?

Let's see what years we have available (Note: Not all years will be available for all measures)

```{r}
unique(gcb_clean$year)
```


*Exercise:* How many unique countries are present in this data frame?

# Time Series Plot

Let's plot territorial emissions for China for all the available years. This is a **time series**: we're showing the change in a single unit (here, a country), over time.

We want to make our code-writing somewhat clever. Rather than inputting "China" everywhere, and then replacing it everywhere if we later want to plot something else, let's define at the top what we'd like:

```{r}
chosen_country <- c("CHN")
chosen_country_name <- "China"
```

These will show up in other code blocks throughout this vignette. Let's also create a data frame that just contains the data for China.

```{r}
data_country <- gcb_clean %>% 
                dplyr::filter(iso3c == chosen_country)
```

## The Pipe Operator %>%

If you're not familiar with the data cleaning and organizing packages included in what's called the `tidyverse`, you might be a little confused by the `%>%` symbol. This is called the "pipe operator": think of it as a funnel.

You take the stuff that came before the pipe, and funnel it through the pipe into the next function that comes below.

In coding terms, what it says is, take the thing that came before (here `gcb_clean`) and insert it into the first argument of the function that comes next (here, `filter`). 

We could also have written 

```{r}
data_country <- dplyr::filter(gcb_clean, iso3c == chosen_country)
```

*Comprehension check:* What has the `filter` function done? How many rows exist in the data frame `data_country`?

Writing `dplyr::filter` just tells R that the function `filter` comes with the `dplyr` package. There are several packages that have a `filter` function, so we specify *which* package's `filter` function we want here. We'll generally state the package for any function that isn't very basic within the vignettes.

Look at the help for this function
```{r}
?filter
```

and click on the `filter` function for `dplyr`.

*Comprehension Check*: Did you see another package that also has a `filter` function? If so, which was it?

You'll see that the first thing that goes into the function is `.data`. `filter` takes a data frame, and keeps only the *rows* for which a condition holds (here, that the iso3c code is the same as we've listed for `chosen_country`.)

The reason for using `%>%` is that there may be many data manipulation operations we'd like to do. Here, it's simple because we just want one thing to happen: we want to keep only the observations for China.

## == or =?

Note the `==` double equals sign is not the same as the single `=` sign. 

The double equals sign means that R is checking whether the condition on the left (here, that `iso3c`) is indeed equal to the thing on the right (here, that it equals `chosen_country`, which is actually `CHN`). 

The single equals sign would be setting the left-hand side equal to the right-hand side. We could have written `chosen_country = c("CHN")` to set the value for `chosen_country`, but in R it's common to use the left-facing arrow `<-` in your scripts, and `=` sign in functions.

So `iso3c == "CHN"` will return `TRUE` if the row is for China, and `FALSE` if the observation is not for China. 


## Filtering with Multiple Conditions

If you click on this data frame, you'll notice a number of missing observations. 

Since we want to plot territorial emissions, we can add a condition to the filter:

```{r}
data_country <- gcb_clean %>% dplyr::filter(iso3c == chosen_country & !is.na(gcb_ghg_territorial))
```

We've added a condition, which is removing the rows for which `gcb_ghg_territorial` is missing. `is.na()` will return `TRUE` if the row is missing (`NA`) in the column `gcb_ghg_territorial`.

The `!` is a negation, so `!is.na()` now returns `TRUE` if the row is *not* missing, so that we keep the rows for which the territorial emissions *are* present.

## Multiple Pipes

For exposition purposes, let's see the `%>%` in action with a little more complexity. Since we're just plotting territorial emissions, let's drop the other columns.

```{r}
data_country <- gcb_clean %>% 
                dplyr::filter(iso3c == chosen_country & !is.na(gcb_ghg_territorial)) %>%
                dplyr::select(year,country_name,iso3c,gcb_ghg_territorial)
```

**Note:** if you're using the pipe, you have to make sure that it goes at the end of the line, not the beginning.

This is saying: first take the data frame `gcb_clean`, and then take only the rows where the `iso3c` is equal to `CHN` and also the `gcb_ghg_territorial` is not missing. Once you've done that, keep only the columns `year`, `country_name`, `iso3c`, `gcb_ghg_territorial`. 

There are lots of ways we could write this. We could split the filter into two separate pipes, we could swap the order of the `filter` and `select`... in this case the order won't matter. But in other cases, with more complicated cleaning, the order may well matter.

*Comprehension check:* Create a new dataframe called `data_country2` that's exactly the same as `data_country` but uses a slightly different way of writing the command to get there from `gcb_clean`.

Now let's get to plotting. The package `ggplot2` is a versatile plotting package that allows you to use a similar syntax for plotting all sorts of figures, from bar charts to complicated maps. There's a [whole ggplot2 book](https://ggplot2-book.org/) that you can use to get into the details, but I've found that economists usually only manipulate a small subset of these options.

We're also going to use the package `ggrepel` to put labels on the graph. The way `ggplot2` typically works is that we start with the data (here, `data_country`), and then we add visual components in layers with each new call to something of the form `geom_xxx`. 

For the purpose of our labels, we're going to select just a few years to label:

```{r}
years_to_show <- c(1958,1959,1960,1961,1978,1995,2000,2015,2019)
```

Now let's generate a plot:

```{r}
my_plot <- ggplot2::ggplot(data = data_country,
               ggplot2::aes(x = year)) +
  ggplot2::geom_point(ggplot2::aes(y =gcb_ghg_territorial, color = "Territorial Emissions")) +
  ggrepel::geom_text_repel(data = base::subset(data_country,year %in%years_to_show), # pick out just these years
                  ggplot2::aes(y = gcb_ghg_territorial,
                      label = year),
                  max.overlaps = 17)+ # a lower number will give fewer total labels; higher will put more labels in
  ggplot2::labs(title = paste0("Territorial Emissions, ",chosen_country_name),
       caption = c("GDP from GCB (2023)"),
       x ="" ,
       y = "Emissions (units here)",
       color = "Emissions" # sets legend name
  )+
  # xlab() +
  # ylab() +
  theme_minimal_plot(#title_size = 20,
             axis_title_x = ggplot2::element_text(color = "black",size = 15),
             axis_title_y = ggplot2::element_text(color = "black", size = 15),
             legend.key = ggplot2::element_rect(fill = "white", # box fill for the legend
                                       colour = "white" # box outlines for the legend
             ),
             legend_position = c(.15,.85) # sets legend position, from [0,1] on X axis then [0,1] on y
  ) +
  #ggplot2::scale_y_continuous(trans = "log10", limits = c(400,100000)) +
  ggplot2::scale_x_continuous(limits = c(1899,2025))

```

You can see the plot in your console by typing `my_plot` in the console.

```{r, echo = FALSE}
url <- "https://github.com/stallman-j/ekonomR/blob/main/output/02_figures/gcb_territorial_emissions_China.png?raw=true"
knitr::include_graphics(url)

```


Now let's save the map, with a function from `ekonomR` that uses the ggplot2's `ggsave` with some simple defaults
```{r}
ggsave_plot(output_folder = here::here("output","02_figures"),
         plotname = my_plot,
         filename = paste0("gcb_territorial_emissions_",chosen_country_name,".png"),
         width = 8,
         height = 6,
         dpi  = 400)
```

# Exercise:

For a country which is  *not* China, plot its greenhouse gas *consumption* (not territorial emissions) over time with the following edits to the code above:

- Edit the y axis labels to contain the appropriate units. They currently say `Emissions (units here)`.
    -*Hint:* You may want to poke around in the [GCB](https://globalcarbonbudget.org/) page to figure out what the units should be.
- Turn the legend off by using `legend_position = "none"` in the correct place in your version of the code chunk above ).
- Correct the caption to contain the correct data attribution. It should be something like "Emissions data from GCB (2023)."
- Comment on these trends relative to what you know about the history of the country. Do you think there's evidence for a Green Kuznets curve for this class? What else would you need to examine in order to make a more definitive claim?


<!-- # Plot Cross Sections -->

<!-- ```{r} -->
<!-- years <- c(1950:2019) # generates a sequence from 1950 to 2019, i.e 1950,1951, ..., 2018,2019 -->
<!-- chosen_years <- c(1950, 2019) # just 1950 and 2019 -->
<!-- ``` -->

<!-- I have a few HEX codes for Yale colors that get output with the following function: -->

<!-- ```{r} -->
<!-- display_hex_colors() -->
<!-- ``` -->

<!-- You'll see four colors outputted. Set a few of them in your local environment with the following code : -->

<!-- ```{r} -->
<!-- yale_lblue     <- "#63aaff" -->
<!-- yale_medblue   <- "#286dc0" -->
<!-- yale_blue      <- "#00356b" -->
<!-- ``` -->

<!-- *Comprehension Exercise:* What's the other color that's available in `display_hex_colors()?` -->

<!-- We're going to manually set the colors for our legend so that they look nice and Yaley. We do this by first defining a vector as below: -->

<!-- ```{r} -->
<!-- colors <- c("Territorial Emissions"    = yale_lblue, -->
<!--             "Consumption Emissions"    = yale_blue, -->
<!--             "Emissions Transfers"      = yale_medblue) -->
<!-- ``` -->

<!-- Let's see how to plot territorial emissions in Sweden for all the available years. -->


<!-- labels_vec <- #unique(data$country_name) # uncomment to try to get all countries -->
<!--   # if you want to use the ISO3 code instead so you can get more labels, use -->
<!--   # unique(data$iso3c) or provide a vector that's like c("USA","CHN","KEN") -->
<!--   c("USA","Sweden","Germany","El Salvador","China","Chile") -->


<!-- #y <- 2019 # uncomment this if you want to examine within the loop to see what's happening -->
<!-- for (y in chosen_years) { -->
<!--   # choose just the data for the current year -->
<!--   data_year_y <- data %>% filter(year == y) -->

<!--   plot <- ggplot(data = data_year_y, -->
<!--                  aes(x = gdp_pc)) + -->
<!--     geom_point(aes(y =le_birth, color = "Life Expectancy at Birth")) + -->
<!--     geom_text_repel(data = subset(data_year_y,country_name %in%labels_vec), # plot just the labels of the countries we want -->
<!--                     aes(y = le_birth, -->
<!--                         label = country_name))+ -->
<!--     geom_point(aes(y =le_15, color = "Life Expectancy at Age 15")) + -->
<!--     geom_text_repel(data = subset(data_year_y,country_name %in%labels_vec), # plot just the labels of the countries we requested -->
<!--                     aes(y = le_15, -->
<!--                         label = country_name))+ -->
<!--     geom_point(aes(y =le_65, color = "Life Expectancy at Age 65")) + -->
<!--     geom_text_repel(data = subset(data_year_y,country_name %in%labels_vec), # plot just the labels of the countries we requested -->
<!--                     aes(y = le_65, -->
<!--                         label = country_name))+ -->
<!--     labs(title = paste0("Life Expectancy at Different Ages and GDP, ",y), -->
<!--          caption = c("GDP from PWT (2022), population data from UN WPP (2022)"), -->
<!--          x ="GDP per capita (units here)" , -->
<!--          y = "Life Expectancy (units)", -->
<!--          color = "" # sets legend name -->
<!--     )+ -->
<!--     # xlab() + -->
<!--     # ylab() + -->
<!--     theme_plot(title_size = 20, -->
<!--                axis_title_x = element_text(color = "black",size = 15), -->
<!--                axis_title_y = element_text(color = "black", size = 15), -->
<!--                legend.key = element_rect(fill = "white", # box fill for the legend -->
<!--                                          colour = "white" # box outlines for the legend -->
<!--                ), -->
<!--                legend.position = c(.15,.85) #"none" # sets legend position, x from [0,1] to y [0,1]. -->
<!--                # remove legend with writing legend.position = "none" instead -->
<!--     ) + -->
<!--     scale_x_continuous(trans = "log10", limits = c(400,100000)) + -->
<!--     scale_y_continuous(limits = c(0,100)) + -->
<!--     scale_color_manual(values = colors) # this sets the legend colors as yale colors -->
<!--   #scale_y_continuous(trans = "log10", limits = c(.05,50)) + -->
<!--   #scale_linetype_manual("",values = c("Predicted Values")) -->

<!--   plot -->

<!--   # I have a save_map and a save_plot function, but the save_map gets used -->
<!--   # more often so it's less buggy at the moment -->
<!--   # good example of "don't let the perfect be the enemy of the `it works by golly I'll take it`" -->

<!--   save_plot(output_folder = file.path(output_figures,"GDP_LE"), -->
<!--            plotname = plot, -->
<!--            filename = paste0("gdp_pc_le_",y,".png"), -->
<!--            width = 9, -->
<!--            height = 6, -->
<!--            dpi  = 400) -->

<!-- } -->

<!-- # Plot One Country Over Time ---- -->

<!-- chosen_country <- c("CHN") -->
<!-- chosen_country_name <- "China" -->

<!-- years_to_show <- c(1958,1959,1960,1961,1978,1995,2000,2015,2019) -->

<!-- # choose just the data for the current year -->
<!-- data_country_c <- data %>% filter(iso3c == chosen_country) -->

<!-- plot <- ggplot(data = data_country_c, -->
<!--                aes(x = gdp_pc)) + -->
<!--   geom_point(aes(y =le_birth, color = "Life Expectancy at Birth")) + -->
<!--   geom_text_repel(data = subset(data_country_c,year %in%years_to_show), # pick out just these years -->
<!--                   aes(y = le_birth, -->
<!--                       label = year), -->
<!--                   max.overlaps = 17)+ # max.overlaps at a lower number will give fewer total labels; higher will put more labels in -->
<!--   geom_point(aes(y =le_15, color = "Life Expectancy at Age 15")) + -->
<!--   geom_text_repel(data = subset(data_country_c,year %in%years_to_show), -->
<!--                   aes(y = le_15, -->
<!--                       label = year), -->
<!--                   max.overlaps = 17)+ -->
<!--   geom_point(aes(y =le_65, color = "Life Expectancy at Age 65")) + -->
<!--   geom_text_repel(data = subset(data_country_c,year %in%years_to_show), -->
<!--                   aes(y = le_65, -->
<!--                       label = year), -->
<!--                   max.overlaps = 17)+ -->
<!--   labs(title = paste0("Life Expectancy and GDP, ",chosen_country_name), # here's another good example of paste0 -->
<!--        # to the rescue. If I wanted to select several countries and try them out before deciding, -->
<!--        # I can just change "chosen_country_name" rather than having to do this all manually -->
<!--        # Or if I wanted, I could make this a loop and loop over a bunch of different countries -->
<!--        # if my interest was in comparing different countries -->
<!--        caption = c("GDP from PWT (2022), population data from UN WPP (2022)"), -->
<!--        x ="GDP per capita (units here)" , -->
<!--        y = "Life Expectancy (units here)", -->
<!--        color = "" # sets legend name -->
<!--   )+ -->
<!--   # xlab() + -->
<!--   # ylab() + -->
<!--   theme_plot(title_size = 20, -->
<!--              axis_title_x = element_text(color = "black",size = 15), -->
<!--              axis_title_y = element_text(color = "black", size = 15), -->
<!--              legend.key = element_rect(fill = "white", # box fill for the legend -->
<!--                                        colour = "white" # box outlines for the legend -->
<!--              ), -->
<!--              legend.position = c(.15,.85) # sets legend position, from [0,1] on X axis then [0,1] on y -->
<!--   ) + -->
<!--   scale_x_continuous(trans = "log10", limits = c(400,100000)) + -->
<!--   scale_y_continuous(limits = c(0,100)) + -->
<!--   scale_color_manual(values = colors) # this sets the legend colors as yale colors -->
<!--   #scale_y_continuous(trans = "log10", limits = c(.05,50)) + # might want this one instead for the PSET -->
<!-- #scale_linetype_manual("",values = c("Predicted Values")) -->


<!-- plot -->

<!-- save_map(output_folder = file.path(output_figures,"GDP_LE"), -->
<!--          plotname = plot, -->
<!--          filename = paste0("gdp_pc_le_",chosen_country_name,".png"), -->
<!--          width = 8, -->
<!--          height = 6, -->
<!--          dpi  = 400) -->




<!-- plot <- ggplot() + -->
<!--   # geom_point(aes(x= gdp_pc, -->
<!--   #               y =le_birth, color = "Data"), -->
<!--   #            data = data) + -->
<!--   geom_point(aes(x = gdp_pc, -->
<!--                 y =fit, color = "Predictions"), -->
<!--             data = predicted_df) + -->
<!--   # why might it look off if we take the actual best-fit line from our regression? -->
<!--   geom_smooth(aes(x = gdp_pc, -->
<!--                y = fit, -->
<!--                color = "Cubic Line of fit"), -->
<!--               data = predicted_df, -->
<!--               formula = y~ x + I(x^2)+I(x^3), -->
<!--               method  = lm)+ -->
<!--   # geom_ribbon(aes(x = gdp_pc, -->
<!--   #                 ymin =ci_low, -->
<!--   #                 ymax = ci_high, -->
<!--   #                 fill = "grey90", -->
<!--   #                 color = "Confidence Bands"), -->
<!--   #             data = predicted_df)+ -->
<!--   labs(title = paste0("Predicted Life Expectancy at Birth and GDP per capita"), -->
<!--        caption = c("GDP from PWT (2022), population data from UN WPP (2022)"), -->
<!--        x ="GDP per capita (units here)" , -->
<!--        y = "Predicated Life Expectancy at Birth (units here)", -->
<!--        color = "" # sets legend name -->
<!--   )+ -->
<!--   # xlab() + -->
<!--   # ylab() + -->
<!--   theme_plot(title_size = 20, -->
<!--              axis_title_x = element_text(color = "black",size = 15), -->
<!--              axis_title_y = element_text(color = "black", size = 15), -->
<!--              legend.key = element_rect(fill = "white", # box fill for the legend -->
<!--                                        colour = "white" # box outlines for the legend -->
<!--              ), -->
<!--              legend.position = c(.15,.85) # sets legend position, from [0,1] on X axis then [0,1] on y -->
<!--   ) + -->
<!--   scale_x_continuous(trans = "log10", limits = c(400,100000)) + -->
<!--   scale_y_continuous(limits = c(0,100)) + -->
<!--   scale_color_manual(values = colors) # this sets the legend colors as yale colors -->
<!-- #scale_y_continuous(trans = "log10", limits = c(.05,50)) + -->
<!-- #scale_linetype_manual("",values = c("Predicted Values")) -->


<!-- plot -->

<!-- save_map(output_folder = file.path(output_figures,"GDP_LE"), -->
<!--          plotname = plot, -->
<!--          filename = paste0("gdp_pc_le_predictions.png"), -->
<!--          width = 8, -->
<!--          height = 6, -->
<!--          dpi  = 400) -->
