---
title: "Basic Randomization: Randomize Presentations"
layout: single
toc: true
toc_label: "Contents"
toc_sticky: true
author_profile: true
date: "2024-10-30"
output: 
  html_document:
    css: mystyle.css
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

library(ekonomR)
library(dplyr)

 # html_document:
  #    css: mystyle.css
  
writeLines("td {border: 1px solid #00356b; padding: 6px;}, th {padding : 12px } th {background-color: #00356b ; color : white; border : 1px solid black; } t d {color : #00356b ; border : 1px solid blue }", con = "mystyle.css")


```

**Make sure** you've got the latest version of `ekonomR`. It's getting updated frequently. 

If you're not sure if your `ekonomR` is up to date or you're new to the woods, you may want to check out the vignette [Getting Started with ekonomR](https://stallman-j.github.io/ekonomR/vignettes/getting-started-with-ekonomR/), and the [Coding Review](https://stallman-j.github.io/ekonomR/vignettes/coding-review/) for a couple key operations I'll be assuming you know.


# Prerequisites

First, bring `ekonomR` into your working library.

``` {r}
library(ekonomR)
```

# Set parameters

We're going to set the path that we want to take the groups from. 

```{r, eval = TRUE, results = FALSE}
data_raw_path <- file.path("P:","2024-2025_academics","ECON-412_2024F","attendance")

filename      <- "ECON-412_2024F_project-groups.xlsx"

in_path       <- file.path(data_raw_path,filename)

group_var     <- "Group"
```

# Bring in groups

Read in the file
```{r}
  groups <- readxl::read_xlsx(path = in_path,
                   col_names = TRUE
  )

num_groups <- length(unique(groups$Group))

paste0("There are ",num_groups," groups.")

just_groups <- groups %>%
               dplyr::group_by(Group) %>%
               dplyr::select(Group) %>%
               dplyr::filter(dplyr::row_number() == 1)
```

Display the groups

```{r}
knitr::kable(groups[,1:2], format="html")
names(groups)
```


# Randomize
Set a "randomization seed"

```{r}
random_seed <- 5 #readline("Pick a positive integer: ") %>% as.numeric()

set.seed(random_seed)
```

Randomly group:

```{r}

order_presentations <- sample(1:num_groups,
                              size = num_groups,
                              replace = FALSE)

just_groups <- cbind(just_groups,order_presentations) %>%
               dplyr::rename(Order = "...2")
```


```{r, cols.print = 4}
groups_with_order <- groups %>%
                     dplyr::left_join(just_groups) %>%
                     dplyr::mutate(Date = ifelse(Order<=7,
                                          yes = Date <- "Monday Nov 4",
                                          no  = Date <- "Wednesday Nov 6")) %>%
                     dplyr::arrange(Order)

paste0("Groups with a number under or equal to ",ceiling(num_groups/2), " will present on Monday.")

knitr::kable(groups_with_order, format="html")

groups_with_rep <- groups_with_order %>%
                   dplyr::group_by(Group)%>%
                   dplyr::filter(dplyr::row_number() == 1) %>%
                   dplyr::rename(Representative = Name)

knitr::kable(groups_with_rep, format="html")


```
