---
title: "Fixed Effects Estimation"
layout: single
toc: true
toc_label: "Contents"
toc_sticky: true
author_profile: true
date: "2024-11-08"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{fixed-effects-estimation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

library(sandwich)
library(modelsummary)

```


**Make sure** you've got the latest version of `ekonomR`. If you've updated the package less recently than Nov 8, 2024, you should install the latest version. You might want to uncomment one or both of the following to run the installation code. (Making sure that the `ekonomR` package is unchecked in your "Packages" tab before you re-install).

```{r, results = FALSE, eval = FALSE} 
#install.packages("remotes")
#remotes::install_github("stallman-j/ekonomR")
```

If you're not sure if your `ekonomR` is up to date or you're new to the woods, you may want to check out the vignette [Getting Started with ekonomR](https://stallman-j.github.io/ekonomR/vignettes/getting-started-with-ekonomR/), and the [Coding Review](https://stallman-j.github.io/ekonomR/vignettes/coding-review/) for a couple key operations I'll be assuming you know.

# Prerequisites

I'm also going to assume that you already know the concepts we discussed in the basic cleaning vignettes, as well as in the [Basic Regression](https://stallman-j.github.io/ekonomR/vignettes/basic-regression/) vignette. This vignette directly follows that.

First, bring `ekonomR` into your working library.

``` {r}
library(ekonomR)
```

# Overview

We're continuing our vignette series of analyzing the evidence of a Green Kuznets curve, or the hypothesis that as a country grows richer, environmental quality first deteriorates and then improves, in the context of greenhouse gas emissions.

In this vignette, we're going to move away from the cross-sectional regression of the [Basic Regression](https://stallman-j.github.io/ekonomR/vignettes/basic-regression/), and focus on controlling for certain types of factors with fixed effects. 

# Data


The dataset we'll be using includes per-capita measures for gross domestic product (GDP) and greenhouse gases (GHGs) from the following sources:

- population data from the [World Population Prospects](https://population.un.org/wpp/Download/Standard/MostUsed/) (WPP)
- GDP data from the [Penn World Tables](https://www.rug.nl/ggdc/productivity/pwt/?lang=en) (PWT)
- greenhouse gas data from the [Global Carbon Budget (GCB)](https://globalcarbonbudget.org/).

We'll bring in the data, and then call `reg_data`` to be this dataset (so that this code is easy to adapt for other datasets), and also filter out observations for which the outcome variable is missing to avoid warnings or errors.
``` {r}
data("ghg_pop_gdp")
# Uncomment below if you'd like to browse the data a little
#names(ghg_pop_gdp)
#View(ghg_pop_gdp)

reg_data <- ghg_pop_gdp %>%
            dplyr::filter(!is.na(gcb_ghg_territorial_pc))
```

# Refresher of fixed effects

If it's been a while since you've thought about fixed effects, there are a few great videos by [Ben Lambert on Youtube](https://www.youtube.com/@SpartacanUsuals) that will get you the intuition. His videos are the first I check out for econometrics concepts I've forgotten.

I recommend the following:

1. [Dummy variable estimators](https://www.youtube.com/watch?v=i7vYh1kCEOY)
2. [Fixed effects estimator](https://www.youtube.com/watch?v=sFvV9b1cGFc)

Another *great* resource for the intuition is [the Chapter on Fixed Effects in *The Effect* by Nick Huntington-Klein](https://theeffectbook.net/ch-FixedEffects.html). It's really approachable.

I'm now going to assume I can use their jargon.

**Rough intuition**

Suppose we're considering just country fixed effects. 

I tend to think of fixed effects regressions as adding a dummy variable for each country, so that each country gets its own Y intercept in a scatter plot where the X axis is the regressor of interest (GDP per capita) and the Y axis is the outcome variable (GHG emissions per capita).

You might also have heard of fixed effects as "de-meaning" or a "within" estimator. 

"De-meaning" is the idea that we can generate a fixed effects estimator by calculating the mean value over all time periods for each country, and then country-by-country subtract that from a regression equation like Equation \ref{eq:eq_1}. 

In that case, then the coefficient of interest $$\beta_1$$ can be described as giving us the change in GHG per capita when we change GDP per capita *relative to the country's mean GDP per capita over all the years we observe it*. 

It's called the "within" estimator because we're now looking *within* a particular country.

These are equivalent, but one way of describing it might stick in your head better.

# Conceptual framework

We're going to continue examining the relationship between greenhouse gases per capita and GDP per capita, but now we'll use the **panel** nature of our data rather than examining **cross sections** at a single point in time.

We have here data from 1950 to 2019, ranging from 55 countries in 1950 to 182 countries in 2019. Observing the same country at multiple times makes this a panel. The fact that not all countries are present in all years makes this panel **unbalanced**.


## Pooled regression

Equation \ref{eq:eq_1} describes a **pooled** panel specification, showing $$\text{GHGpc}$$, greenhouse gas emissions per capita in country $$i$$ and pooling all the years $$t$$, as a function of $$\text{GDPpc}$$, per-capita GDP. 

We're including the cubic and quadratic here so that we can have nonlinear effects.

$$\text{GHGpc}_{i,t} = \beta_0 + \beta_1 \text{GDPpc}_{i,t} + \beta_2 \text{GDPpc}_{i,t}^2 + \beta_3 \text{GDPpc}_{i,t}^3 + \varepsilon_{i,t}\label{eq:eq_1}$$

It's pooled because we've thrown in all the years and countries without explicitly considering that countries are going to exhibit persistence across time, and that particular years are going to affect all countries.

The comparison we're making here is a *between* variation: that is, we're comparing *between* countries and *between* years. The variation in GHG per capita comes from comparing the United States in 1960 to Kenya in 2019.

## Time fixed effects

Equation \ref{eq:eq_2} adds in year fixed effects, denoted by $$\alpha_t$$.

$$\text{GHGpc}_{i,t} = \beta_1 \text{GDPpc}_{i,t} + \beta_2 \text{GDPpc}_{i,t}^2 + \beta_3 \text{GDPpc}_{i,t}^3 + \alpha_t + \varepsilon_{i,t}\label{eq:eq_2}$$


You can think of time fixed effects as adding a dummy variable that shifts the *average* level of greenhouse gases per capita for each year, but that this average is *constant across all countries*. 

For instance, we would expect that greenhouse gases are rising globally over time, because countries have generally been emitting more over time. 

However, even in more recent years, the financial recession of 2008 might have been associated with relatively smaller average global emissions, because many economies were seeing stagnating output.

The other way of thinking about it explicitly as a *within* estimator is to say that now we're looking at the variation *between* countries, but *within* the same period of time. 



## Entity fixed effects

Equation \ref{eq:eq_3} adds in entity (here, country) fixed effects, denoted by $$\alpha_i$$.

$$GHGpc_{i,t} = \beta_1 \text{GDPpc}_{i,t} + \beta_2 \text{GDPpc}_{i,t}^2 + \beta_3 \text{GDPpc}_{i,t}^3 + \alpha_i + \varepsilon_{i,t}\label{eq:eq_3}$$


We can think of country fixed effects as adding a dummy variable for *each* country that is *constant across all years*. 

This reflects, for instance, that Sweden has generally been quite environment-conscious so would tend to emit less regardless of how it is doing economically; that the United States is a rich country and tends to emit quite a lot of greenhouse gases per capita; and that Brazil has and has had a large Amazon rainforest which soaks up a large amount of carbon dioxide. 

This is the *within* variation by country. If we're looking at *within* variation in GDP per capita, it's that we're comparing how Ecuador did economically in 2019 relative to how Ecuador did economically in 1983, given that it's Ecuador.

## Two-way fixed effects {#twfe}

Equation \ref{eq:eq_4} includes both country fixed effects, denoted by $$\alpha_i$$, as well as year fixed effects, denoted by $$\alpha_t$$. 

$$GHGpc_{i,t} = \beta_1 \text{GDPpc}_{i,t} + \beta_2 \text{GDPpc}_{i,t}^2 + \beta_3 \text{GDPpc}_{i,t}^3+ \alpha_i + \alpha_t + \varepsilon_{i,t}\label{eq:eq_4}$$


Fixed effects can become quite unmanageable quite quickly if we keep thinking of the "adding an intercept" interpretation. We now have 70 years and 182 countries. Compare that with just three other regressors.

This is precisely why statistics packages in practice compute using some form of de-meaning rather than adding in hundreds, thousands, or more intercepts.

Two-way fixed effects are trickier to interpret, but the key is to think of every set of fixed effects as adding another *within*.

We've subtracted out the country-level means as well as the year-level means. What we're left with is how countries vary from year to year relative to that annual average level across all countries, and then again relative to their own country-average value.

In other words, the variation that we have left after we've included both sets of fixed effects is the variation relative to the average for the country and in that particular year.




# Prep regressions

I'm assuming you're familiar with how I like to set up regression tables from the [Basic Regression](https://stallman-j.github.io/ekonomR/vignettes/basic-regression/) vignette, so let's jump in.

We'll still use the territorial greenhouse gases per capita.

We'll add another element to the lists, `fevars`, where we'll store which variables we're taking fixed effects over. `iso3c` is the three-letter code for a country, so it's equivalent to using country fixed effects.

**Choose regression variables**

Pick out our regression variables.

```{r}
reg_1_vars <- list(outvar = "gcb_ghg_territorial_pc",
                   regvars = c("gdp000_pc","I(gdp000_pc^2)","I(gdp000_pc^3)"),
                   fevars  = NULL)

reg_2_vars <- list(outvar = "gcb_ghg_territorial_pc",
                   regvars = c("gdp000_pc","I(gdp000_pc^2)","I(gdp000_pc^3)"),
                   fevars = "year")

reg_3_vars <- list(outvar = "gcb_ghg_territorial_pc",
                   regvars = c("gdp000_pc","I(gdp000_pc^2)","I(gdp000_pc^3)"),
                   fevars = c("iso3c"))

reg_4_vars <- list(outvar = "gcb_ghg_territorial_pc",
                   regvars = c("gdp000_pc","I(gdp000_pc^2)","I(gdp000_pc^3)"),
                   fevars = c("year","iso3c"))
```

**Make regression formulas**

Now that we've chosen our variables, we can make our regression formulas:

```{r}
reg_eq_1 <- ekonomR::reg_equation(outcome_var    = reg_1_vars$outvar,
                                  regressor_vars = reg_1_vars$regvars,
                                  fe_vars        = reg_1_vars$fevars)

reg_eq_2 <- ekonomR::reg_equation(outcome_var    = reg_2_vars$outvar,
                                  regressor_vars = reg_2_vars$regvars,
                                  fe_vars        = reg_2_vars$fevars)

reg_eq_3 <- ekonomR::reg_equation(outcome_var    = reg_3_vars$outvar,
                                  regressor_vars = reg_3_vars$regvars,
                                  fe_vars        = reg_3_vars$fevars)

reg_eq_4 <- ekonomR::reg_equation(outcome_var    = reg_4_vars$outvar,
                                  regressor_vars = reg_4_vars$regvars,
                                  fe_vars        = reg_4_vars$fevars)
```

**Choose standard errors**

`ekonomR` has another convenience function called `cluster_formula()` that takes in either a character vector of the variables you want to cluster or a regression formula, and tries to intelligently pull out the variables you used as fixed effects and turn them into a formula that you can then put into `modelsummary` or `feols` as the variables you should cluster by, so that you don't have to hard-code everything. 

Let's set those up, as well, then. We call them `vcov_x` because they'll go into the `vcov` (variance-covariance matrix) argument. 


```{r}

vcov_1 <- ekonomR::cluster_formula(reg_eq_1)
vcov_2 <- ekonomR::cluster_formula(reg_eq_2)
vcov_3 <- ekonomR::cluster_formula(reg_eq_3)
vcov_4 <- ekonomR::cluster_formula(reg_eq_4)
```

You should get a warning message from the first one, because `reg_eq_1` is just `gcb_ghg_territorial_pc ~ gdp000_pc + I(gdp000_pc^2) + I(gdp000_pc^3)` and there are no fixed effects in that regression. That's fine. It outputs `"HC1"` which will give us robust standard errors anyways in `modelsummary`.

These are good defaults, but you could obviously change them manually if you had a reason to.

Just like we're going to do with the model, it would be helpful to aggregate these options into a list.

```{r}
vcov_list <- list(vcov_1,
                  vcov_2,
                  vcov_3,
                  vcov_4)
```

# Generate models
Let's make up our models. The first is just using `lm()` as before. The rest will have to use `feols::fixest()` to get the right estimation. With `feols` we can get clustered standard errors right away.

```{r}
model_1 <- lm(reg_eq_1, data = reg_data)
model_2 <- fixest::feols(reg_eq_2, 
                         vcov = vcov_2,
                         data = reg_data)
model_3 <- fixest::feols(reg_eq_3, 
                         vcov = vcov_3,
                         data = reg_data)
model_4 <- fixest::feols(reg_eq_4, 
                         vcov = vcov_4,
                         data = reg_data)
```

We can look at some of these outputs. Uncomment the below if you'd like to check them out manually.

```{r}
#summary(model_1)
#summary(model_2)
#summary(model_3)
#summary(model_4)
```

Put these models into a handy list

```{r}
models <- list(
  "(1)" = model_1,
  "(2)" = model_2,
  "(3)" = model_3,
  "(4)" = model_4)
```

# Generate output table

This should be starting to feel familiar. First we'll choose the title and notes, figure out how to add the dependent variable means, and produce our final output.

**Choose title and notes**

```{r}
my_title <- paste0("GHG and GDP per capita relationship \\label{tab:fe_reg}")

table_notes <- "Robust standard errors given in parentheses in column (1). The other columns have standard errors clustered at the units for which fixed effects are calculated. Population data are obtained from UN-DESA (2023). Gross domestic product (GDP) in 2017 chained PPP thousand USD per capita (PWT 2023). Greenhouse gases in tonnes of carbon per year from GCB (2024)."

cov_labels <- c("Intercept","GDP pc","(GDP pc)$^2$", "(GDP pc)$^3$")
```


**Generate dependent variable means**

Let's again generate the mean of the dependent variable and decide the place to put it into the table.

This time, however, we're also going to say whether the *column* has certain fixed effects attached to it. We do this by expanding the `rows` dataframe to include a few more rows.

We also need to add to the `attribute` of rows to place all three rows, otherwise the rows with fixed effects get shunted to the bottom.

``` {r}
n_total_regvars <- length(unique(c(reg_1_vars$regvars,reg_2_vars$regvars,reg_3_vars$regvars,reg_4_vars$regvars)))


rows <- data.frame("term" = c("Mean","Year FE","Country FE"),
                   "(1)"  = c(round(mean(reg_data[[reg_1_vars$outvar]]),2), "N","N"),
                   "(2)"  = c(round(mean(reg_data[[reg_2_vars$outvar]]),2), "Y","N"),
                   "(3)"  = c(round(mean(reg_data[[reg_3_vars$outvar]]),2), "N","Y"),
                   "(4)"  = c(round(mean(reg_data[[reg_4_vars$outvar]]),2), "Y","Y"))

attr(rows, 'position') <- c(2*n_total_regvars+4,2*n_total_regvars+5,2*n_total_regvars+6)            

```

This is now what the rows will look like. 

If you're getting `NA` in the rows for a particular variable when you do your own analysis, make sure you remove your missing values of the outcome variable, or examine the `na.rm` in the help for `?mean`, because the mean will default to missing if any values are missing.


| term      | X.1 | X.2 | X.3 | X.4 |
|------     |-----|-----|-----|-----|
| Mean      | 1.29| 1.29| 1.29| 1.29|
| Year FE   | N   | Y   | N   | Y   |
| Country FE| N   | N   | Y   | N   |

**Produce final output**

We have a few more things to remove, so `gof_omit` grew a little bit. We also added `vcov_list` to tell `modelsummary` which standard errors to pick, and just for illustration I've set `fmt = 6` to make the coefficients show out to 6 decimal places.

```{r}

my_table <- modelsummary::modelsummary(models,
                                       stars = FALSE,
                                       vcov = vcov_list,
                                       fmt  = 6, # to 4 decimal places
                                       coef_rename = cov_labels,
                                       title = my_title,
                                       format = "latex",
                                       add_rows = rows,
                                       gof_omit = "AIC|BIC|RMSE|Log.Lik|Std.Errors|FE:|Adj.|F",
                                       escape = FALSE,
                                       notes = table_notes
)   %>%
  tinytable::group_tt(j = list("GHGpc" =2:5))
  
```


```{r, eval = FALSE, echo = TRUE}
my_table
```

```{r, eval = TRUE, echo = FALSE}

modelsummary::modelsummary(models,
             stars = FALSE,
             vcov = vcov_list,
             fmt = 6,
             coef_rename = cov_labels,
             title = my_title,
             output = "kableExtra",
             add_rows = rows,
             gof_omit = "AIC|BIC|RMSE|Log.Lik|Std.Errors|FE:|Adj.|F",
             escape = FALSE,
             notes = table_notes

) %>%
  kableExtra::add_header_above(c(" "=1, "GHGpc" =4))

            
```

**Save**

```{r, eval = FALSE}
  tinytable::save_tt(my_table,
                     output = here::here("output","01_tables","fixed_effects_table.tex"),
                     overwrite = TRUE)
  
   tinytable::save_tt(my_table,
                     output = here::here("output","01_tables","fixed_effects_table.docx"),
                     overwrite = TRUE)
```
If you get an output error like `cannot open the connection`, make sure the file is closed on your local computer before running the command. R can't overwrite the file if it's open.


```{r, eval = TRUE, echo = FALSE}
  tinytable::save_tt(my_table,
                     output = here::here("example-project","output","01_tables","fixed_effects_table.tex"),
                     overwrite = TRUE)
  
   tinytable::save_tt(my_table,
                     output = here::here("example-project","output","01_tables","fixed_effects_table.docx"),
                     overwrite = TRUE)
```

Do **not** copy and paste latex output directly into LaTex or Overleaf. There's a much nicer way to do it.

If you're using Overleaf, make sure you have the following code in your preamble:

``` {r, eval = FALSE}
\usepackage{tabularray}
\usepackage{float}
\usepackage{graphicx}
\usepackage{rotating}
\usepackage[normalem]{ulem}
\UseTblrLibrary{booktabs}
\NewTableCommand{\tinytableDefineColor}[3]{\definecolor{#1}{#2}{#3}}
\newcommand{\tinytableTabularrayUnderline}[1]{\underline{#1}}
\newcommand{\tinytableTabularrayStrikeout}[1]{\sout{#1}}
```

The code `tinytable::save_tt()` will have given you a `.tex` file.

Now, you can drag and drop the entire table `.tex` file into, preferably, a folder in your Overleaf project called something like `Tables`. 

In the [Overleaf ECON 412 folder](https://www.overleaf.com/read/wsrdjdckwmbz#f4467b), for instance, I've uploaded the table from this exercises into the path `tables/tables-from-r/fixed_effects_table.tex`. 

When you update your table, you just have to update the drag and drop.

If you'd like to be even fancier, you can sync your Overleaf with Dropbox or Github so that the tables update automatically when you run your R code, and then again when you compile in Overleaf.

Because we set a label above, we can also cross-reference the table. An example of how to do this is given in the file `templates/examples/inserting-tables.tex`.

# Default output with `modelsummary_reg_default`.

The function `modelsummary_reg_default` might be helpful for generating nice looking regression tables, but this is an optional section.

`ekonomR`'s function `modelsummary_reg_default()` includes all the above in a simple regression format with nice defaults. It works for both linear models and fixed effects. It hasn't been modified to get fancier than that, or to handle fancy interaction terms in the regression formulas.

All you need to do is be able to input the regression variables in a list such that the function can read it, and you'll get a pretty reasonable output table for your fixed effects regression of choice. 

Create the regression variables. You can have however many regressions you want, and here's an example with 3.

```{r}
reg_1_vars <- list(outvar = "gcb_ghg_territorial_pc",
                   regvars = c("gdp000_pc","I(gdp000_pc^2)","I(gdp000_pc^3)"),
                   fevars  = NULL)

reg_2_vars <- list(outvar = "gcb_ghg_territorial_pc",
                   regvars = c("gdp000_pc","I(gdp000_pc^2)","I(gdp000_pc^3)"),
                   fevars = "year")

reg_3_vars <- list(outvar = "gcb_ghg_territorial_pc",
                   regvars = c("gdp000_pc","I(gdp000_pc^2)","I(gdp000_pc^3)"),
                   fevars = c("iso3c"))

reg_vars_list <- list(reg_1_vars,
                      reg_2_vars,
                      reg_3_vars)
```

Run `modelsummary_reg_default()` with default settings. This will export a file by default, and also push to your console a summary of all the models you generated. 

```{r, eval = TRUE}
my_table <- modelsummary_reg_default(reg_vars_list = reg_vars_list,
                         data = reg_data)

```

We can see this output in Console and the viewer

```{r, eval = FALSE}
my_table 
```

Let's see a little more flexibility.

You can see the defaults and get help with `?modelsummary_reg_default`.

`modelsummary()` is already an extremely flexible package, as is `fixest`. The point of this function `modelsummary_reg_default` is not to replace the flexibility of these packages, but to provide a set of simple defaults that get your output up and running without much effort.

```{r, eval = TRUE}
my_table <- modelsummary_reg_default(reg_vars_list = reg_vars_list,
                         data = reg_data,
                         my_title = "This is my title \\label{tab:my_table_label}",
                         table_notes = list(a = "Top level notes",
                                            b = "Second level notes"),
                         cov_labels = c("Intercept","GDP pc","(GDP pc)$^2$", "(GDP pc)$^3$"),
                         fe_names = c("Year FE","Country FE"),
                         depvar_means = c(1.23,4.59,"lego"),
                         fmt = 6,
                         export_output = FALSE)

my_table <- my_table %>%
  tinytable::group_tt(j = list("GHGpc" =2:4))


# generate output folder if it doesn't already exist

if (!dir.exists(here::here("output","01_tables"))) dir.create(here::here("output","01_tables"), recursive = TRUE)

 tinytable::save_tt(my_table,
                     output = here::here("output","01_tables","fixed_effects_table.tex"),
                     overwrite = TRUE)

  tinytable::save_tt(my_table,
                     output = here::here("output","01_tables","fixed_effects_table.docx"),
                     overwrite = TRUE)
```


# Exercises

1. What would the intuitive relationship between greenhouse gases per capita and GDP per capita be in Equation \ref{eq:eq_1} if $$\beta_1$$ were positive, $$\beta_2$$ were negative, and $$\beta_3$$ were effectively zero? (Think of what the relationship would be at low levels of GDP per capita, intermediate levels, and then high levels of GDP per capita)

    - What about if $$\beta_1$$ were negative, $$\beta_2$$ were positive, and $$\beta_3$$ were negative?
    - (**Hint:** you might find it helpful to explore some toy numbers in an app like [Desmos](https://www.desmos.com/calculator).)
    
2. Generate a new model, called `model_5`, which is the same as `model_4` but instead clustered by just `iso3c` instead of both `year` *and* `iso3c`. What's the standard error on `gdp000_pc` in the old `model_4`, and in this new `model_5`? 

    - You don't need to make up a new table to do this; you can just figure it out from your R console output.

3. Reproduce the final table, but instead of using *per capita* variables use just GDP and just greenhouse gas emissions by country. Remember to change the column title, the table title, the table notes, and the covariate labels.

    - Briefly describe which table, using the per capita variables or the raw country-level variables, that you find more convincing if we're trying to evaluate the relationship between economic growth and greenhouse gas emissions. Why do you prefer the one you prefer?

# Just the code, please

```{r, eval = FALSE}

# install if you need it
#install.packages("remotes")
#remotes::install_github("stallman-j/ekonomR")


# bring in ekonomR

library(ekonomR)

# bring in the data
data("ghg_pop_gdp")
# Uncomment below if you'd like to browse the data a little
#names(ghg_pop_gdp)
#View(ghg_pop_gdp)

# remove non-missing values for now
reg_data <- ghg_pop_gdp %>%
  dplyr::filter(!is.na(gcb_ghg_territorial_pc))

# choose regression variables

reg_1_vars <- list(outvar = "gcb_ghg_territorial_pc",
                   regvars = c("gdp000_pc","I(gdp000_pc^2)","I(gdp000_pc^3)"),
                   fevars  = NULL)

reg_2_vars <- list(outvar = "gcb_ghg_territorial_pc",
                   regvars = c("gdp000_pc","I(gdp000_pc^2)","I(gdp000_pc^3)"),
                   fevars = "year")

reg_3_vars <- list(outvar = "gcb_ghg_territorial_pc",
                   regvars = c("gdp000_pc","I(gdp000_pc^2)","I(gdp000_pc^3)"),
                   fevars = c("iso3c"))

reg_4_vars <- list(outvar = "gcb_ghg_territorial_pc",
                   regvars = c("gdp000_pc","I(gdp000_pc^2)","I(gdp000_pc^3)"),
                   fevars = c("year","iso3c"))

# create regression equations

reg_eq_1 <- ekonomR::reg_equation(outcome_var    = reg_1_vars$outvar,
                                  regressor_vars = reg_1_vars$regvars,
                                  fe_vars        = reg_1_vars$fevars)

reg_eq_2 <- ekonomR::reg_equation(outcome_var    = reg_2_vars$outvar,
                                  regressor_vars = reg_2_vars$regvars,
                                  fe_vars        = reg_2_vars$fevars)

reg_eq_3 <- ekonomR::reg_equation(outcome_var    = reg_3_vars$outvar,
                                  regressor_vars = reg_3_vars$regvars,
                                  fe_vars        = reg_3_vars$fevars)

reg_eq_4 <- ekonomR::reg_equation(outcome_var    = reg_4_vars$outvar,
                                  regressor_vars = reg_4_vars$regvars,
                                  fe_vars        = reg_4_vars$fevars)

# examine the equations if you like

#reg_eq_1
#reg_eq_2
#reg_eq_3
#reg_eq_4

# choose standard errors

vcov_1 <- ekonomR::cluster_formula(reg_eq_1)
vcov_2 <- ekonomR::cluster_formula(reg_eq_2)
vcov_3 <- ekonomR::cluster_formula(reg_eq_3)
vcov_4 <- ekonomR::cluster_formula(reg_eq_4)

# put standard error choices together in a list

vcov_list <- list(vcov_1,
                  vcov_2,
                  vcov_3,
                  vcov_4)

# generate models
model_1 <- lm(reg_eq_1, data = reg_data)
model_2 <- fixest::feols(reg_eq_2, 
                         vcov = vcov_2,
                         data = reg_data)
model_3 <- fixest::feols(reg_eq_3, 
                         vcov = vcov_3,
                         data = reg_data)
model_4 <- fixest::feols(reg_eq_4, 
                         vcov = vcov_4,
                         data = reg_data)

# look at outputs
#summary(model_1)
#summary(model_2)
#summary(model_3)
#summary(model_4)

# put models into a list
models <- list(
  "(1)" = model_1,
  "(2)" = model_2,
  "(3)" = model_3,
  "(4)" = model_4)

# choose title and notes 
my_title <- paste0("GHG and GDP per capita relationship \\label{tab:fe_reg}")

table_notes <- "Robust standard errors given in parentheses in column (1). The other columns have standard errors clustered at the units for which fixed effects are calculated. Population data are obtained from UN-DESA (2023). Gross domestic product (GDP) in 2017 chained PPP thousand USD per capita (PWT 2023). Greenhouse gases in tonnes of carbon per year from GCB (2024)."

cov_labels <- c("Intercept","GDP pc","(GDP pc)$^2$", "(GDP pc)$^3$")

# generate dependent variable means
n_total_regvars <- length(unique(c(reg_1_vars$regvars,reg_2_vars$regvars,reg_3_vars$regvars,reg_4_vars$regvars)))


rows <- data.frame("term" = c("Mean","Year FE","Country FE"),
                   "(1)"  = c(round(mean(reg_data[[reg_1_vars$outvar]]),2), "N","N"),
                   "(2)"  = c(round(mean(reg_data[[reg_2_vars$outvar]]),2), "Y","N"),
                   "(3)"  = c(round(mean(reg_data[[reg_3_vars$outvar]]),2), "N","Y"),
                   "(4)"  = c(round(mean(reg_data[[reg_4_vars$outvar]]),2), "Y","Y"))

attr(rows, 'position') <- c(2*n_total_regvars+4,2*n_total_regvars+5,2*n_total_regvars+6)            


# produce final output

my_table <- modelsummary::modelsummary(models,
                                       stars = FALSE,
                                       vcov = vcov_list,
                                       fmt  = 6, # to 4 decimal places
                                       coef_rename = cov_labels,
                                       title = my_title,
                                       format = "latex",
                                       add_rows = rows,
                                       gof_omit = "AIC|BIC|RMSE|Log.Lik|Std.Errors|FE:|Adj.|F",
                                       escape = FALSE,
                                       notes = table_notes
)   %>%
  tinytable::group_tt(j = list("GHGpc" =2:5))


# generate output folder if it doesn't already exist
if (!dir.exists(here::here("output","01_tables"))) dir.create(here::here("output","01_tables"), recursive = TRUE)


# save 
tinytable::save_tt(my_table,
                   output = here::here("output","01_tables","fixed_effects_table.tex"),
                   overwrite = TRUE)

tinytable::save_tt(my_table,
                   output = here::here("output","01_tables","fixed_effects_table.docx"),
                   overwrite = TRUE)

# use modelsummary_reg_default() instead

# generate lists of regression variables
reg_1_vars <- list(outvar = "gcb_ghg_territorial_pc",
                   regvars = c("gdp000_pc","I(gdp000_pc^2)","I(gdp000_pc^3)"),
                   fevars  = NULL)

reg_2_vars <- list(outvar = "gcb_ghg_territorial_pc",
                   regvars = c("gdp000_pc","I(gdp000_pc^2)","I(gdp000_pc^3)"),
                   fevars = "year")

reg_3_vars <- list(outvar = "gcb_ghg_territorial_pc",
                   regvars = c("gdp000_pc","I(gdp000_pc^2)","I(gdp000_pc^3)"),
                   fevars = c("iso3c"))

reg_vars_list <- list(reg_1_vars,
                      reg_2_vars,
                      reg_3_vars)

# generate output
my_table <- modelsummary_reg_default(reg_vars_list = reg_vars_list,
                         data = ghg_pop_gdp,
                         my_title = "This is my title \\label{tab:my_table_label}",
                         table_notes = list(a = "Top level notes",
                                            b = "Second level notes"),
                         cov_labels = c("Intercept","GDP pc","(GDP pc)$^2$", "(GDP pc)$^3$"),
                         fe_names = c("Year FE","Country FE"),
                         depvar_means = c(1.23,4.59,8.90,"lego"),
                         fmt = 6,
                         export_output = FALSE)

my_table <- my_table %>%
  tinytable::group_tt(j = list("GHGpc" =2:4,"GHGpc" = 5))

# generate output folder if it doesn't already exist

if (!dir.exists(here::here("output","01_tables"))) dir.create(here::here("output","01_tables"), recursive = TRUE)

 tinytable::save_tt(my_table,
                     output = here::here("output","01_tables","fixed_effects_table.tex"),
                     overwrite = TRUE)
```




