---
title: "Getting Started with ekonomR"
layout: single
toc: true
toc_label: "Contents"
toc_sticky: true
author_profile: true
date: "2024-09-23"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{getting-started-with-ekonomR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Getting Started

If you're just starting your project, you can use `ekonomR` to create your project folders so that you have a tidy structure that can scale up and stay organized while you and your collaborators are working and adding files to it, or even adding collaborators.

The folder structure that `ekonomR` creates has been adapted from the recommendations made by the [World Bank Development Impact Group (DIME) ](https://dimewiki.worldbank.org/Data_Management). DIME has built out a Wiki covering best practices for project structuring for impact evaluations of randomized controlled trials or policy changes. These impact evaluations tend to be large, complicated, and expensive projects with many contributors that change often (e.g. with adding or removing interns).

Even if you don't have multiple collaborators or a large-budget operation, their setup is helpful for making your project clear and replicable, whether to your collaborators or just your future self.

Unfortunately for R users, DIME's tools are targeted towards Stata users. This perhaps begs the question, why use R at all?

## Why R at All?

Over the past decade or so, the fact of R being an open-source software with many contributors has allowed it to surpass Stata's capabilities in many areas, particularly in frontier applications like spatial analysis and network theory. 

R users frequently make their own packages, which you can typically install either from [CRAN](https://cran.r-project.org/) (which requires that the developers follow a strict set of guidelines to make their packages usable across many platforms over time); or from Github (where the standards are laxer, but the innovation faster). I think essentially having two sources, one heavily structured and the other highly flexible, is an advantage of R over Python, where package sourcing often causes conflicts.

R and Python seem to have largely converged in terms of usability and capability. Some users gravitate towards Python, so build out better functionality with better help and documentation there. Web scraping, as well as many [API calls](https://www.cloudflare.com/learning/security/api/what-is-api-call/), seem moderately more fleshed out with Python. Other users, including statisticians, have migrated towards R, so for instance R's data-plotting capabilities are quite advanced.

Both R and Python are able to implement code from the other language. R's [`reticulate`](https://rstudio.github.io/reticulate/) package allows you either to translate Python to R and back, or just call Python from within your R project. (To do: link to a vignette that uses a Python API call within R).

## What's a package?

`ekonomR` is an R package. A package in R world is a bundle of a few things:

1. Functions: **functions** take inputs, do something to them, and produce an output. A math professor of mine is fond of saying, "Functions describe the world!"
2. Documentation
3. (Possibly) example data
4. (Hopefully) vignettes providing examples of the package in use.

My imagined view of how packages come about is the following: some developer spends quite a bit of time writing code to fulfill a particular purpose. 

Then, or simultaneously, if the developer is clever, this developer muses, "Golly, this coding was a fair bit of hassle. But I can imagine other contexts in which I or others would like to use this type of code. Wouldn't it be nice if I could re-use what they wrote without having to manually adjust everything?"

That's exactly what a package does.^[You don't have to be a hardcore developer to build an R package, though. If you're familiar with R, it's really not hard. If you ever get to the point where you find yourself repeating very similar code in your scripts and having to scrounge around in your old files, you might want to consider trying to make your own package! [Here's](https://tinyheero.github.io/jekyll/update/2015/07/26/making-your-first-R-package.html) a great place to start if you decide to make that leap.]

R packages are very varied in their capabilities and scope.

For example, the R package [`comtradr`](https://docs.ropensci.org/comtradr/index.html) allows you to access the API for the [UN-Comtrade database](https://comtradeplus.un.org/), a primary source of trade data.

The R package [`causalToolbox`](https://github.com/forestry-labs/causalToolbox), on the other hand, allows you to use frontier machine-learning techniques to estimate how the effects of a treatment can vary across the population that was treated.

The [`fishR` website](https://fishr-core-team.github.io/fishR/pages/packages.html) gives an example of the functions of a variety of fish-related R packages.

# Getting the right R

Sometimes functions stop working because you're using an older version of R. If you skip this step and use an outdated R, you might spend hours or days trying to fix your code when all you really needed was ten minutes to update R.

If you last used R more than three months ago, uninstall R and R studio on your device and then reinstall from [here](https://posit.co/download/rstudio-desktop/). 

First install R, and then install R Studio.

Once you've done that, click on your RStudio icon to open a new session.

# Choosing your home folder

Now decide what folder you'd like your project to live in. We'll create project folders in this home folder. 

If you're coming from ECON 412, I highly recommend putting your project in your OneDrive, since Yale has free storage in there and you'll be able to collaborate with your group by syncing your folders remotely. 

Unless your project is going to be data intensive, it's likely best that your data, code and output live all in the same place, which `ekonomR` calls the `home_folder`.

Log into your OneDrive (after installing if need be) and find the folder path to your OneDrive folder. Mine is `J - Yale University`, for instance. You might also use GoogleDrive or Dropbox, but be cautioned that your teammates may not have this software.

## What if I'm not married to a particular cloud?

If you're setting out into the brave new world of research and trying to decide what cloud to use (and you *absolutely* should be using one), I recommend considering [pCloud](https://www.pcloud.com/), which offers lifetime plans and similar features to Dropbox. A 10 TB plan, for instance, beats Dropbox for price competitivity within 5 years if you get it on sale (and sales are frequent. Just sign up for the newsletter).

If you're using big data, it's useful that pCloud only draws storage from the person who's the host of the data. Dropbox, on the other hand, counts storage against every user.

# Create a New RStudio Project

In your RStudio session, click `File -> New Project`. 

Choose the option `New Directory`, and then `New Project`. Creating an R project saves a bunch of your settings so that it's less hassle to use R, and lets you pull up your project files and folders easily.

In `Directory name:`, write something like `ECON-412_your-initials`. When you ultimately share this project, you don't want everyone having the same name or you might have file path conflicts.

Click `Browse` and navigate to your OneDrive (or other folder on your device where you've decided your folder is living). A good place to put this would be under a file path that looks like `Your Name - Yale University/Projects`. You might have to create the `Projects` subfolder.

Then click `Create Project`.

This should open your project in R.

If you created the project and successfully browsed to the right folder, you should be able to access this file path with `getwd()`, which returns the file path of your **w**orking **d**irectory.^[Think of a directory as just a folder. There are differences, but I'll almost always use the terms interchangeably.] 

Type in your console the following:

``` r
my_home_folder <- getwd()

```

To see the path in full, type in the console:

``` {r, results = FALSE, eval = FALSE}
getwd()

```
You can do this more cleverly with the [`here` package](https://here.r-lib.org/), which tries to figure out where your home folder is and creates paths relative to this folder. `ekonomR` loads the `here` package when it's loaded, so you should be able to input this command directly:

``` {r, results = FALSE, eval = FALSE}
here::here()
```

This should give you the same result as what you'd done with `getwd()`. You could therefore set 

```{r, results = FALSE, eval = FALSE}
my_home_folder <- here::here()
```

Using the `here()` function is a more robust way to write your home folder than using `getwd()`, and it's portable across platforms (for instance if you have a PC and your collaborator has a Mac).


You should probably save a shortcut to this folder somewhere that is easy to visualize, since you'll be clicking into this folder often. If you have a PC, right-clicking on the folder once you've navigated to it and then clicking `Pin to Quick Access` is a good way to make it easily accessible. Mac probably has something similar.

# Installing ekonomR

Let's install this `ekonomR` package. 


## First-time Installation

If this is the first time you've installed `ekonomR`, you need to first install the package `remotes` from CRAN that allows you to install a package from GitHub (which is where `ekonomR` lives).

Installing a package is like buying an app off an app store.

You can do both of these with the following code:
``` {r, results = FALSE, eval = FALSE}
install.packages("remotes")
remotes::install_github("stallman-j/ekonomR")
```

Then you'll want to bring the package into your library. Putting a package in your library is like downloading an app that you've already bought onto a new device.

``` {r, results = FALSE}
library(ekonomR)
```

## Re-installing 

If you've installed `ekonomR` before, you should re-install the package again so that you can access the (frequent) updates. Go into RStudio in the `Packages` tab (which is in the same window as the tabs for `Files`, `Plots`, `Help`, and `Viewer`). Scroll down to the `ekonomR` row and make sure that the check box is **unchecked**. This will ensure that the re-installation doesn't conflict with files that currently exist.

Then, do the following:

``` {r, results = FALSE, eval = FALSE}
remotes::install_github("stallman-j/ekonomR")
library(ekonomR)
```


Now you can access the functions of the `ekonomR` package.

# Create your project folders

If you're making a new project, you can use the `create_folders` function from `ekonomR` to create your project folders on your local device. 

In the `my_home_folder` folder that you set, `create_folders` is, unsurprisingly, going to create a bunch of folders.

To see the help documentation for this function, put the following in your RStudio console:

``` {r, results = FALSE, eval = FALSE}
?create_folders

```

Putting a question mark before the function name for any function accesses the function description and help files. If you get an error to this command, it's probably the case that the package that contains the function is either not installed on your computer, not loaded into your library, or both.

To create your project folders, type the following into your RStudio console.

``` {r, results = FALSE, eval = FALSE}
create_folders(home_folder = my_home_folder)

```

You should open up the folder `my_home_folder` and see what's been created. You should have a folder for code, a folder for data, a folder for output, and a few other collaboration folders, along with a `.Rproj` file that's storing your project settings.

This is the entire folder structure that gets created on your computer.

```bash
. (home_folder)
├── articles
├── citations
├── documents
├── code
│   └── 00_functions
│   └── 01_download
│   └── 02_cleaning
│   └── 03_analysis
│   └── 04_plots
│   └── 05_simulations
│   └── scratch
├── output
│   └── 01_tables
│   └── 02_figures
│   └── 03_maps
│   └── x_manual-output
│   └── scratch
├── data
│   └── 00_manually-downloaded
│   └── 01_raw
│   └── 02_temp
│   └── 03_clean
├── your-project-name.Rproj

. data_external_folder (optional file path)
├── data
│   └── 00_manually-downloaded
│   └── 01_raw
│   └── 02_temp
│   └── 03_clean

. data_pii_folder (optional file path)
├── data
│   └── 00_manually-downloaded
│   └── 01_raw
│   └── 02_temp
│   └── 03_clean
```

## Other Options

You can set `output` and `data` to live in a different folder than your `home_folder`. This may be a good option if you generate a lot of plots that take up space, or if you have a lot of data and just want to sync your code to GitHub. In this case, set a separate path for `output_folder` or `data_folder` If you don't set these file paths separately, they'll get created within `home_folder`.

If you have sensitive information, you can also set `data_pii_folder` so that private data lives separately. If you have an external hard drive that, say, holds big data, you can set this with `data_external_folder`. If you don't set these, these folders don't get created.

Instead of `my_home_folder` which we accessed as the working directory with `getwd()`, you can also set the file path manually by finding what folder on your device you'd like to use, and getting its full path.

The `file.path()` function in R allows you to ignore whether you're on a Mac or Windows (which have different backslash and forward slash conventions). So `J - Yale University/Projects` becomes `file.path("J - Yale University","Projects")`.

In that case, you would input something like the following in your console:

``` {r, results = FALSE, eval = FALSE}
create_folders(home_folder = file.path("J - Yale University","Projects","ECON-412_js"))

```

If the folder `ECON-412_js` doesn't already exist, the `create_folders` function will create it.

Now you're ready to go with your project! 


# A Typical Workflow

Here's the way a typical project analysis should go and how the folder structure you've just created works with that. 

In the simplest case, the paths mentioned below are all taken relative to the project folder you set as `home_folder`. For example, if your R Project is stored in `C:/Projects/ECON-412_js`, then when you see `data/01_raw` you should think of this as meaning the file path `C:/Projects/ECON-412_js/data/01_raw`.

[Getting Started with ekonomR](https://stallman-j.github.io/ekonomR/vignettes/getting-started-with-ekonomR/) discusses more complicated cases.


## Decide your research question

This is not trivial! In fact, it can be very difficult. A research question is a question that you try to answer through your analysis of the data and/or creation of an economics model. 

A good way to see examples of research questions is to examine the abstracts of economics papers you've read. You should be able to determine the research question that the paper is trying to answer in any economics paper abstract.
  
## Decide your emipirical strategy
  
Are you running a regression and trying to identify a causal effect? If so, what's the equation of the regression you're running? What's your mode of identification? (usually: regression discontinuity, difference-in-difference analysis, or instrumental variable regression)

Are you creating a model? What data are you hoping your model will match?

Are you trying to measure or account for something? How will you know if your accounting is accurate?

## Find your data

Your research question and empirical strategy will get revised throughout your project, particularly as you settle on your data.

That's okay! The research process is not linear.

See my [recommended data sources](https://stallman-j.github.io/how-tos/data) page to get started on finding data.

## Access your data 

### Intermediate R Users
If you're used to using R already, challenge yourself to download the data entirely using R. This package has a few functions for making that easier. (Vignette coming soon).

You should tell your files to download into `data/01_raw`.

The R code for downloading the files should live in `code/01_download`.

### Early R Users

If you're new to R, just download your data manually, for instance by navigating in your web browser to the download page and clicking on a "Download" button. In that case, save your data into 
`data/00_manually downloaded` so that your collaborators know that they shouldn't be looking for a downloading script in the folder `code/01_download`.
  
## Clean your data

The R code for cleaning your data should live in `code/02_cleaning`. It's good practice to number your cleaning scripts to delineate the order in which they should be run. (To include: vignette on cleaning and merging).

Your data should get pulled from `data/01_raw` (if you downloaded your data using an R script) or `data/00_manually-downloaded` (if you clicked something like a `Download` button on a website somewhere). 

It should be sent to either `data/02_temp` or `data/03_clean`


**How do I know whether to put data in `02_temp` or `03_clean`?**

If your code takes a long time to run, you might want to run a cleaning script on a smaller portion of your data, or you might want to break up your cleaning into steps. If you're running into trouble cleaning, it can be useful to create intermediate data files. A temp data file is anything that you're not going to run analysis on or make figures or maps out of, but that you might want to bring into R to take a look at.

In short: if you're not running analysis on it but still want to create the files, put it in `data/02_temp`. If you're creating data that you can analyze directly, put it directly into `03_clean`.

**How do I know if my data is cleaned?**

In most cases, you'll know that your data counts as cleaned once you have done the following:

- Set up your data in long format: this means that an observation (a row) signifies a unique combination of:
  
  - a unit (e.g. a country, person, firm, city)
  - (possibly) a time period (e.g. month, year, quarter)
  
  - Usually you can tell you're there when you have one column that contains your units (say, countries) and another column that tracks the time period.
  
  - If your rows contain countries and your columns contain years, for instance, your data is not ready for analysis and will need to be transformed from what's called "wide" format into "long" format. Vignettes with examples will be put up to show you simple ways to achieve this.

- Accounted for oddities in the data
  
  - This typically includes analyzing the data's missing values, correcting entries that may have been erroneous, and getting a sense of how well the data were collected and inputted
  
- Merged the datasets that you're planning to use into a single cleaned dataset.
  
  - Oftentimes we end up bringing multiple datasets together. For instance, an analysis of the relationship between greenhouse gases per capita and incomes per capita would require data on greenhouse gases, data on incomes, and data on populations
  
  - In this example you might have a cleaning script to clean each of the datasets on greenhouse gases, incomes, and population, as well as a final fourth script merging these three (cleaned) datasets together. (Vignette showing this to be added).

  
- Defined all the variables you're planning to use. 
 
  - Broadly speaking, you **define a variable** whenever you create a column (variable) that didn't come with your raw data. This includes summing two variables that already exist, taking log transformations of a variable, truncating variables above or below certain values, squaring, and so forth.

  - You should *not* be defining variables in plotting, simulation, or analysis scripts. In other words, the number of *columns* in your data should not increase in any of your analysis or plot scripts.
  
    - **Note:** Your final merging script is where you should define your variables, if at all possible. That way, when you look back because you forgot how you created a particular variable that is not present in the raw data's documentation, you know where to find it: it's in your merge script. Otherwise you have to look through all your cleaning scripts, and your collaborators also have to look through all your cleaning scripts, in order to find *anything*, and you and your collaborators might end up defining variables that are similar but not equal in different scripts, and then you have a royal mess.
    
    - If you can't manage to define all your variables in your final merge script for some reason, then define those variables at the end of the cleaning script that sends a particular dataset to `data/03_clean`. Then you, future self, and other collaborators won't have too many places where you have to look.


## Run your analysis and make your plots

Once you have cleaned data that lives in `data/03_clean`, you're ready to do analysis.

Analysis scripts should live in `code/03_analysis`. They should take data from `data/03_clean`, and produce outputs that go to one of `output/01_tables`, `output/02_figures`, or `output/03_maps`, depending on the type of output that gets produced.

Similarly, plotting and mapping scripts should live in `code/04_plots`, take data from `data/03_clean`, and produce outputs that go to one of `output/02_figures`, or `output/03_maps`.

Occasionally you might run simulations, in which case the code should live in `code/05_simulations` that uses clean data from `data/03_clean`, and sends output into the `output` folder.

If you have scratch code that's a hodgepodge, put it in `code/scratch` while you're toying around, and then split up the script into its appropriate folders once you're a little further along.

If you're working with LaTex/Overleaf, you'll upload the figures, maps and tables into your Overleaf files, and these will get put into your final documents.

## Communicate with your audience

If you're working with LaTeX/Overleaf (in which case you should check out [LaTekonomer](https://stallman-j.github.io/LaTekonomer/) if you're doing a big project, and the [ECON 412 LaTex Guide](https://stallman-j.github.io/LaTekonomer/how-tos/ECON-412_overleaf/) if you're in ECON 412), your documents will be produced in Overleaf.

When you generate PDFs or presentations, save them into `documents`.

## Cite your sources

[Zotero](https://www.zotero.org/) is a lovely reference manager that I personally find superior to the competition. It integrates with LaTeX; Microsoft Word; Google Docs; and a bunch of other sources.

It works seamlessly with Firefox and several other web browsers, so that it easily downloads the citations when you're at the article page and integrates them into your library.

If you have more than ten sources and you're planning on writing more than three more papers which require references, it's worthwhile to invest in Zotero. 

There are ways to set it up so that it's more convenient that what you get out of the box. You can set it to download articles into PDFs that can live in your cloud (so that you can access the PDFS on any device), so that you can read any article that you've ever found the citation for. You can also set it to generate citation keys so that you can cite sources in LaTex very easily.

(I have a video tutorial for managing these settings. Let me know if you'd like me to make a guide.)

If you're using a citation manager like Zotero with a software like LaTex/Overleaf, you'll generate a `.bib` file that contains your references. The `citations` folder is a great place to store this file.

## Save Important Articles

If you have important articles to save, put them in `articles`. If you're working with Zotero, though, you can access these at any time through Zotero instead.

# Next Steps

If you'd like to walk through how `ekonomR` works through a whole project sequence, check out the [Vignettes](https://stallman-j.github.io/ekonomR/vignettes/vignettes/)).

A good place to get a feel for how `ekonomR` tries to get you thinking about the research process is [Simple Plotting with a Review of R Coding Basics](https://stallman-j.github.io/ekonomR/vignettes/basic-plotting/).

Each vignette is self-contained, however, if you're looking for something in particular.
