---
title: "modelsummary_reg_default"
layout: single
toc: true
toc_label: "Contents"
toc_sticky: true
author_profile: true
date: "2024-11-08"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{modelsummary-reg-default}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Default output with `modelsummary_reg_default`.

The function `modelsummary_reg_default` might be helpful for generating nice looking regression tables, but this is an optional section.

`ekonomR`'s function `modelsummary_reg_default()` includes all the above in a simple regression format with nice defaults. It works for both linear models and fixed effects. It hasn't been modified to get fancier than that, or to handle fancy interaction terms in the regression formulas.

All you need to do is be able to input the regression variables in a list such that the function can read it, and you'll get a pretty reasonable output table for your fixed effects regression of choice. 

Create the regression variables. You can have however many regressions you want, and here's an example with 3.

```{r}
reg_1_vars <- list(outvar = "gcb_ghg_territorial_pc",
                   regvars = c("gdp000_pc","I(gdp000_pc^2)","I(gdp000_pc^3)"),
                   fevars  = NULL)

reg_2_vars <- list(outvar = "gcb_ghg_territorial_pc",
                   regvars = c("gdp000_pc","I(gdp000_pc^2)","I(gdp000_pc^3)"),
                   fevars = "year")

reg_3_vars <- list(outvar = "gcb_ghg_territorial_pc",
                   regvars = c("gdp000_pc","I(gdp000_pc^2)","I(gdp000_pc^3)"),
                   fevars = c("iso3c"))

reg_vars_list <- list(reg_1_vars,
                      reg_2_vars,
                      reg_3_vars)
```

Run `modelsummary_reg_default()` with default settings. This will export a file by default, and also push to your console a summary of all the models you generated. 

```{r, eval = TRUE}
my_table <- modelsummary_reg_default(reg_vars_list = reg_vars_list,
                         data = reg_data)

```

We can see this output in Console and the viewer

```{r, eval = FALSE}
my_table 
```

Let's see a little more flexibility.

You can see the defaults and get help with `?modelsummary_reg_default`.

`modelsummary()` is already an extremely flexible package, as is `fixest`. The point of this function `modelsummary_reg_default` is not to replace the flexibility of these packages, but to provide a set of simple defaults that get your output up and running without much effort.

```{r, eval = TRUE}
my_table <- modelsummary_reg_default(reg_vars_list = reg_vars_list,
                         data = reg_data,
                         my_title = "This is my title \\label{tab:my_table_label}",
                         table_notes = list(a = "Top level notes",
                                            b = "Second level notes"),
                         cov_labels = c("Intercept","GDP pc","(GDP pc)$^2$", "(GDP pc)$^3$"),
                         fe_names = c("Year FE","Country FE"),
                         depvar_means = c(1.23,4.59,"lego"),
                         fmt = 6,
                         export_output = FALSE)

my_table <- my_table %>%
  tinytable::group_tt(j = list("GHGpc" =2:4))


# generate output folder if it doesn't already exist

if (!dir.exists(here::here("output","01_tables"))) dir.create(here::here("output","01_tables"), recursive = TRUE)

 tinytable::save_tt(my_table,
                     output = here::here("output","01_tables","fixed_effects_table.tex"),
                     overwrite = TRUE)

  tinytable::save_tt(my_table,
                     output = here::here("output","01_tables","fixed_effects_table.docx"),
                     overwrite = TRUE)
```
